---
title: "Basics of R expressions and conditionals"
subtitle: "Stat 133, Fall 2017"
author: "Gaston Sanchez"
output: github_document
fontsize: 11pt
urlcolor: blue
---

> ### Learning Objectives:
>
> - Understand the concept of R Expressions
> - Learn the difference between simple and compound expressions
> - Understand the use of braces to group expressions
> - Conditional structures: if-then-else, and `switch()`

------

## R Expressions

Before talking about conditional structures and loops, we must first talk
about __expressions__.

R programs are made up of expressions. These can be either _simple_ expressions 
or _compound_ expressions. Compound expressions consist of simple expressions 
separated by semicolons or newlines, and grouped within braces.

```r
# structure of a compound expression
# with simple expressions separated by semicolons
{expression_1; expression_2; ...; expression_n}

# structure of a compound expression
# with simple expressions separated by newlines
{
  expression_1
  expression_2
  expression_n
}
```

Here's a less abstract example:

```{r}
# simple expressions separated by semicolons
{"first"; 1; 2; 3; "last"}

# simple expressions separated by newlines
{
  "first"
  1
  2
  3
  "last"
}
```

Writing compound expressions like those in the previous example is not 
something common among R users. Although the expressions are perfectly valid,
these examples are very dummy (just for illustration purposes).

I discourage you from grouping multiple expressions with semicolons because
it makes it difficult to inspect things. As for the expressions separated by 
newlines, they do play an important role but they are typically used together 
with other programming structures (e.g. functions, conditionals, loops).


### Every expression has a value

A fundamental notion about expressions is that 
__every expression in R has a value__. If you have a simple expression like:

```{r}
a <- 5
```

then `a` has the value 5. In contrast, the value of a compound 
expression is the value of the last evaluated expression.

Here's one example of a compund expression. Note that the entire expression is
assigned to `x`:

```{r}
x <- {5; 10}
```

- What is the value of `x`? 
    + 5?
    + 10?
    + 5, 10?

In this case, the compound expression has a value of `10`, which was the last
expression inside the braces.

Here's the same compound expression as above, but now with simple expressions
in newlines:

```{r}
x <- {
  5
  10
}

x
```

To make sure you don't forget it, repeat this mantra:

> - Every expression in R has a value: the value of the last evaluated statement.
> - Every expression in R has a value: the value of the last evaluated statement.
> - Every expression in R has a value: the value of the last evaluated statement.

---

### Assignments within Compound Expressions

It is possible to have assignments within compound expressions and the 
values of the variables which this produces can be used in later expressions.

```{r}
# simple expressions (made up of assignments) separated by newlines
{
  one <- 1
  pie <- pi
  zee <- "z"
}

one
pie
zee
```

Here's another example:

```{r}
z <- { x = 10 ; y = x^2; x + y }

x
y
z
```

Now that we've introduced the concept of R expressions, let's introduce 
conditional structures

---

## Conditionals

Every programming language comes with a set of structures that allows us to
have control over how commands are executed. One of these structures is called
__conditionals__, and as its name indicates, they are used to evaluate 
conditions.

The most common conditional structure, conceptually speaking, is the 
__if-then-else__ statement. This type of statement makes it possible to choose 
between two (possibly compound) expressions depending on the value of a (logical) condition.

In R (as in many other languages) the if-then-else statement has the following
structure:

```r
if (condition) {
  # do something
} else {
  # do something else
}
```

As you can tell, the `if` clause works like a function: `if(condition)`.
Likewise, braces are used to group one or more expressions. If the condition 
to be evaluated is true, then just the expressions inside the first pair of
braces are executed. If the condition is false, then the expressions inside
the second pair of braces are executed:

```{r}
x <- 1

if (x > 0) {
  print("positive")
} else {
  print("not positive")
}
```

For readability reasons, most users prefer to write `if (condition)` instead of 
`if(condition)`. The _condition_ is an expression that when evaluated returns
a __logical__ value of length one. In other words, whatever you pass as the 
input of the `if` clause, it has to be something that becomes `TRUE` or `FALSE`

```{r}
if (TRUE) {
  print("TRUE")
} else {
  print("FALSE")
}
```

```{r}
if (FALSE) {
  print("TRUE")
} else {
  print("FALSE")
}
```


#### Minimalist If-then-else

`if` statements can be written in different forms, depending on the types of 
expressions that are evaluated. If the expressions of both the _True_ part and 
the _False_ part are simple expressions, the if-then-else can be simplified as:

```r
if (condition) expression_1 else expression_2
```

With simple expressions there's actually no need to use braces:

```{r if1}
x <- 10
if (x > 0) y <- sqrt(x) else y <- -sqrt(-x)

y
```

The previous statement can be written more succinctly in R as:

```{r if2}
x <- 10
y <- if (x > 0) sqrt(x) else -sqrt(-x)

y
```

Again, even though the previous commands are perfectly OK, it is preferred to
use braces when working with conditional structures. This is a good practice 
that improves readibility:

```{r}
# embrace braces: use them as much as possible!
x <- 10
if (x > 0) {
  y <- sqrt(x) 
} else {
  y <- -sqrt(-x)
}

y
```


### Simple If's

There is a simplified form of if-then-else statement which is available when 
there is no expression in the False part to evaluate. This statement has the general form:

```r
if (condition) expression
```

and is equivalent to:

```r
if (condition) expression else NULL
```

Here's an example:

```{r simple_if}
x <- 4
y <- 2
if (x > y) print("x is greater than y")
```


### Multiple If's

A common situation involves working with multiple conditions at the same time. 
You can chain multiple if-else statements:

```{r}
y <- 1 # Change this value!

if (y > 0) {
  print("positive")
} else if (y < 0) {
  print("negative")
} else {
  print("zero?")
}
```


#### Your turn!

Write R code that will "squish" a number into the interval [0, 100], so that a 
number less than 0 is replaced by 0 and a number greater than 100 is replaced 
by 100.

```{r}
z <- 100 * pi

# Fill in the following if-else statements. You may (or may not) 
# have to add or subtract else if or else statements.
if (TRUE) { # Replace TRUE with a condition.
  
} else if (TRUE) { # Replace TRUE with a condition.
  
} else {
  
}
```


### Switch

Working with multiple chained if's becomes cumbersome. Consider the following
example that uses several if's to convert a day of the week into a number:

```{r}
# Convert the day of the week into a number.
day <- "Tuesday" # Change this value!

if (day == 'Sunday') {
  num_day <- 1
} else {
  if (day == "Monday") {
    num_day <- 2
  } else {
    if (day == "Tuesday") {
      num_day <- 3
    } else {
      if (day == "Wednesday") {
        num_day <- 4
      } else {
        if (day == "Thursday") {
          num_day <- 5
        } else {
          if (day == "Friday") {
            num_day <- 6
          } else {
            if (day == "Saturday") {
              num_day <- 7
            }
          }
        }
      }
    }
  }
}


num_day
```

Working with several nested if's like in the example above can be a nightmare.

In R, you can get rid of many of the braces like this:

```{r}
# Convert the day of the week into a number.
day <- "Tuesday" # Change this value!

if (day == 'Sunday') {
  num_day <- 1
} else if (day == "Monday") {
  num_day <- 2
} else if (day == "Tuesday") {
  num_day <- 3
} else if (day == "Wednesday") {
  num_day <- 4
} else if (day == "Thursday") {
  num_day <- 5
} else if (day == "Friday") {
  num_day <- 6
} else if (day == "Saturday") {
  num_day <- 7
}

num_day
```

But still we have too many if's, and there's a lot of repetition in the code.
If you find yourself using many if-else statements with identical structure for 
slightly different cases, you may want to consider a __switch__ statement 
instead:

```{r}
# Convert the day of the week into a number.
day <- "Tuesday" # Change this value!

switch(day, # The expression to be evaluated.
  Sunday = 1,
  Monday = 2,
  Tuesday = 3,
  Wednesday = 4,
  Thursday = 5,
  Friday = 6,
  Saturday = 7,
  NA) # an (optional) default value if there are no matches
```

Switch statements can also accept integer arguments, which will act as indices 
to choose a corresponding element:

```{r}
# Convert a number into a day of the week.
day_num <- 3 # Change this value!

switch(day_num,
  "Sunday",
  "Monday",
  "Tuesday",
  "Wednesday",
  "Thursday",
  "Friday",
  "Saturday")
```


### Function `stop()`

Consider the formula for the area $A$ of a circle of given radius $r$:

$$
A = \pi r^2
$$

We can write a function `circle_area()` that calculates the area of a circle. 
This function takes one argument `radius`, and we can give `radius` a default 
value of 1:

```{r}
circle_area <- function(radius = 1) {
  pi * radius^2
}
```

For example:

```{r}
# default (radius 1)
circle_area()

# radius 3
circle_area(radius = 3)
```

What happens if you give radius a negative value?

```{r}
# default (radius 1)
circle_area(-1)
```

The function `circle_area()` still returns a value even when the radius is
negative, which does not make sense. In many cases, it is desirable to make
some checkings on the provided input. If the input is not appropriate, we
should stop the execution of the function. For this purpose, we can use the
function `stop()`

```{r, eval = FALSE}
circle_area <- function(radius = 1) {
  if (radius < 0) {
    stop("radius must be positive")
  }
  pi * radius^2
}

# this should work
circle_area(1)

# this should not work
circle_area(-1)
```

