---
title: "Random Numbers and Simulations"
subtitle: "Stat 133, Fall 2017"
author: "Gaston Sanchez"
output: github_document
fontsize: 11pt
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE)
```

> ### Learning Objectives
>
> - How to use R to simulate chance processes
> - Getting to know the function `sample()`
> - Simulate flipping a coin
> - Visualize relative frequencies

------

## Introduction

Random numbers have many applications in science and computer programming,
especially when there are significant uncertainties in a phenomenon of interest.
In this tutorial we'll look at a basic problem that involves working
with random numbers and creating simulations.

More specifically, let's see how to use R to simulate basic chance processes 
like tossing a coin.


## Let's flip a coin

Chance processes, also referred to as chance experiments, have to do with 
actions in which the resulting outcome turns out to be different in each 
occurrence.

Typical examples of basic chance processes are tossing one or more coins, 
rolling one or more dice, selecting one or more cards from a deck of cards, 
and in general, things that can be framed in terms of drawing tickets out of 
a box (or any other type of container: bag, urn, etc.).

You can use your computer, and R in particular, to simulate chances processes.
In order to do that, the first step consists of learning how to create a 
virtual coin, or die, or box with tickets.


### Creating a coin

The simplest way to create a coin with two sides, `"heads"` and `"tails"`, is 
with an R vector via the _combine_ function `c()`

```{r}
coin <- c("heads", "tails")
```

You can also create a _numeric_ coin that shows `1` and `0` instead of
`"heads"` and `"tails"`:

```{r}
num_coin <- c(0, 1)
```

Likewise, you can also create a _logical_ coin that shows `TRUE` and `FALSE` 
instead of `"heads"` and `"tails"`:

```{r}
log_coin <- c(TRUE, FALSE)
```


## Tossing a coin

Once you have an object that represents the _coin_, the next step involves 
learning how to simulate tossing the coin. One way to simulate the action of 
tossing a coin in R is with the function `sample()` which lets you draw 
random samples, with or without replacement, from an input vector. 

To toss the coin use `sample()` like this:

```{r}
coin <- c('heads', 'tails')

# toss the coin
sample(coin)
```

The call to `sample()` is equivalent to:

```{r}
sample(coin, size = 1)
```

with the argument `size = `, specifying that we want to take a sample of size 1
from the input vector `coin`.

The function `sample()` is one of the functions that allow you to generate 
random numbers. R has a familiy of functions that generates random numbers
following a certain distribution. You can find more information by checking
the help documentation of `?Distributions`


### Function `sample.int()`

Another function related to `sample()` is `sample.int()` which simulates 
drawing random integers. The main argument is `n`, which represents the maximum 
integer to sample from: `1, 2, 3, ..., n`

```{r}
sample.int(10)
```


### Random Samples

By default, `sample()` draws each element in `coin` with the same probability. 
In other words, each element is assigned the same probability of being chosen. 
Another default behavior of `sample()` is to take a sample of the specified 
`size` __without replacement__. If `size = 1`, it does not really matter whether 
the sample is done with or without replacement. 

To draw two elements WITHOUT replacement, use `sample()` like this:

```{r}
# draw 2 elements without replacement
sample(coin, size = 2)
```

What if we try to toss the coin three or four times?

```{r}
# trying to toss coin 3 times
sample(coin, size = 3)
```

Notice that R produced an error message. This is because the default behavior 
of `sample()` cannot draw more elements that the length of the input vector.

To be able to draw more elements, we need to sample WITH replacement, which is
done by specifying the argument `replace = TRUE`, like this:

```{r}
# draw 4 elements with replacement
sample(coin, size = 4, replace = TRUE)
```



## The Random Seed

The way `sample()` works is by taking a random sample from the input vector. 
This means that every time you invoke `sample()` you will likely get a different 
output.

In order to make the examples replicable (so you can get the same output as me),
you need to specify what is called a __random seed__. This is done with the 
function `set.seed()`. By setting a _seed_, every time you use one of the random 
generator functions, like `sample()`, you will get the same values.

```{r}
# set random seed
set.seed(1257)

# toss a coin with replacement
sample(coin, size = 4, replace = TRUE)
```

All computations of random numbers are based on deterministic algorithms, so 
the sequence of numbers is not truly random. However, the sequence of numbers
appears to lack any systematic pattern, and we can therefore regard the 
numbers as random.

Every time you use one of the random generator functions in R, the call 
produces different numbers. For replication and debugging purposes, it is 
useful to get the same sequence of random numebrs every time we run the script.
This functionality is obtained by setting a __seed__ before we start generating
the numebrs. The seed is an integer and set by the function `set.seed()`

```{r}
set.seed(123)
runif(4)
```

If we set the seed to `123` again, the sequence of uniform random numbers is
regenerated:

```{r}
set.seed(123)
runif(4)
```

If we don't specify a seed, the random generator functions set a seed based
on the current time. That is, the seed will be different each time we run the 
script and consequently the sequence of random numbers will also be different.


## Sampling with different probabilities

Last but not least, `sample()` comes with the argument `prob` which allows you 
to provide specific probabilities for each element in the input vector.

By default, `prob = NULL`, which means that every element has the same 
probability of being drawn. In the example of tossing a coin, the command 
`sample(coin)` is equivalent to `sample(coin, prob = c(0.5, 0.5))`. In the 
latter case we explicitly specify a probability of 50% chance of heads, and 
50% chance of tails:

```{r echo = FALSE}
# tossing a fair coin
coin <- c("heads", "tails")

sample(coin)
sample(coin, prob = c(0.5, 0.5))
```

However, you can provide different probabilities for each of the elements in 
the input vector. For instance, to simulate a __loaded__ coin with chance of 
heads 20%, and chance of tails 80%, set `prob = c(0.2, 0.8)` like so:

```{r}
# tossing a loaded coin (20% heads, 80% tails)
sample(coin, size = 5, replace = TRUE, prob = c(0.2, 0.8))
```


-----


## Simulating tossing a coin

Now that we have all the elements to toss a coin with R, let's simulate flipping 
a coin 100 times, and use the function `table()` to count the resulting number 
of `"heads"` and `"tails"`:

```{r}
# number of flips
num_flips <- 100

# flips simulation
coin <- c('heads', 'tails')
flips <- sample(coin, size = num_flips, replace = TRUE)

# number of heads and tails
freqs <- table(flips)
freqs
```

In my case, I got `r freqs[1]` heads and `r freqs[2]` tails. Your results will 
probably be different than mine. Some of you will get more `"heads"`, some of 
you will get more `"tails"`, and some will get exactly 50 `"heads"` and 50 
`"tails"`.

Run another series of 100 flips, and find the frequency of `"heads"` and `"tails"`:

```{r}
# one more 100 flips
flips <- sample(coin, size = num_flips, replace = TRUE)
freqs <- table(flips)
freqs
```


## Tossing function

Let's make things a little bit more complex but also more interesting.
Instead of calling `sample()` every time we want to toss a coin, we can 
write a `toss()` function:

```{r}
# x: coin object (a vector)
# times: number of tosses
toss <- function(x, times = 1) {
  sample(x, size = times, replace = TRUE)
}

# basic call
toss(coin)

# toss 5 times
toss(coin, 5)
```

We can make the function more versatile by adding a `prob` argument that let
us specify different probabilities for `heads` and `tails`

```{r}
# x: coin object (a vector)
# times: number of tosses
toss <- function(x, times = 1, prob = NULL) {
  sample(x, size = times, replace = TRUE, prob = prob)
}

# toss a loaded coin 10 times
toss(coin, times = 10, prob = c(0.8, 0.2))
```


## Counting Frequencies

The next step is to toss a coin several times, and count the frequency of 
`heads` and `tails`

```{r}
tosses <- toss(coin, times = 100)
table(tosses)
```

We can also count the relative frequencies:

```{r}
table(tosses) / length(tosses)
```

To make things more interesting, let's consider of the frequency of `heads`
evolves of a series of `n` tosses.

```{r}
n <- 500
tosses <- toss(coin, times = n)
heads_freq <- cumsum(tosses == 'heads') / 1:n
```

In this case, we can make a plot of the relative frequencies:

```{r head_freqs_plot}
plot(heads_freq, type = 'l', lwd = 2, col = 'tomato', las = 1,
     ylim = c(0, 1))
abline(h = 0.5, col = 'gray50')
```
