---
title: "Lab 5: First contact with dplyr and ggplot2"
subtitle: "Stat 133, Fall 2017"
author: "Gaston Sanchez"
output: github_document
urlcolor: blue
---

> ### Learning Objectives:
>
> - Get started with `"dplyr"`
> - Get to know the basic dplyr verbs:
> - `slice()`, `filter()`, `select()`
> - `mutate()`
> - `arrange()`
> - `summarise()`
> - `group_by()`
> - Get started with `"ggplot2"`
> - Produce basic plots with `ggplot()`

------

## Manipulating and Visualizing Data Frames

Last week you started to manipulate data tables 
(under the class of `"data.frame"` objects) using bracket notation, 
`dat[ , ]`, and the dollar operator, `dat$name`, in order to select specific 
rows, columns, or cells. In addition, you have been creating charts with 
functions like `plot()`, `boxplot()`, and `barplot()`, which are part of the
`"graphics"` package.

In this lab, you will start learning about other approaches to manipulate tables
and create statistical charts. We are going to use the functionality of the 
package `"dplyr"` to work with tabular data in a more consistent way. This is a 
fairly recent package introduced a couple of years ago, but it is based on more
than a decade of research and work lead by Hadley Wickham.

Likewise, to create graphics in a more consistent and visually pleasing way, 
we are going to use the package `"ggplot2"`, also originally authored by 
Hadley Wickham, and developed as part of his PhD more than a decade ago.

Use the first hour of the lab to get as far as possible with the material 
associated to `"dplyr"`. Then use the second hour of the lab to work on 
graphics with `"ggplot2"` (all the material is in this file).

While you follow this lab, you may want to open these cheat sheets:

- [dplyr cheatsheet](../cheat-sheets/data-transformation-cheatsheet.pdf)
- [ggplot2 cheatsheet](../cheat-sheets/ggplot2-cheatsheet-2.1.pdf)


### Installing packages

I'm assuming that you already installed the packages `"dplyr"` and `"ggplot2"`. 
If that's not the case then run on the console the command below 
(do NOT include this in your `Rmd`):

```r
# don't worry too much if you get a warning message
install.packages(c("dplyr", "ggplot2"))
```

Remember that you only need to install a package once! After a package has been 
installed in your machine, there is no need to call `install.packages()` again on 
the same package. What you should always invoke in order to use the functions 
in a package is the `library()` function:

```{r warning=FALSE, message=FALSE}
# don't forget to load the packages
library(dplyr)
library(ggplot2)
```

__About loading packages:__ Another rule to keep in mind is to always load any 
required packages at the very top of your script files (`.R` or `.Rmd` or `.Rnw` files). 
Avoid calling the `library()` function in the middle of a script. Instead, 
load all the packages before anything else.


-----

## NBA Players Data

```{r nba, echo = FALSE}
dat <- read.csv('../data/nba2017-players.csv', stringsAsFactors = FALSE)
```

The data file for this lab is the same you used last week: `nba2017-players.csv`, 
whic is located in the `data/` folder of the course github repository. 
I assume that you already downloaded a copy of the csv file to your computer. 
If that is not the case, here's one option to get your own copy:

```r
# download RData file into your working directory
github <- "https://github.com/ucb-stat133/stat133-fall-2017/raw/master/"
csv <- "data/nba2017-players.csv"
download.file(url = paste0(github, csv), destfile = 'nba2017-players.csv')
```

To import the data in R you can use the base function `read.csv()`, or you 
can also use `read_csv()` from the package `"readr"`:

```{r read, eval = FALSE}
# with "base" read.csv()
dat <- read.csv('nba2017-players.csv', stringsAsFactors = FALSE)

# with "readr" read_csv()
dat <- read_csv('nba2017-players.csv')
```


-----

## Basic `"dplyr"` verbs

To make the learning process of `"dplyr"` gentler, Hadley Wickham proposes 
beginning with a set of five _basic verbs_ or operations for data frames 
(each verb corresponds to a function in `"dplyr"`):

- __filter__: keep rows matching criteria
- __select__: pick columns by name
- __mutate__: add new variables
- __arrange__: reorder rows
- __summarise__: reduce variables to values

I've modified Hadley's list of verbs a little bit: 

- `filter()`, `slice()`, and `select()`: subsetting and selecting rows and columns
- `mutate()`: add new variables
- `arrange()`: reorder rows
- `summarise()`: reduce variables to values
- `group_by()`: grouped (aggregate) operations

-----

## Filtering, slicing, and selecting

`slice()` allows you to select rows by position:

```{r}
# first three rows
three_rows <- slice(dat, 1:3)
three_rows
```

`filter()` allows you to select rows by condition:

```{r}
# subset rows given a condition
# (height greater than 85 inches)
gt_85 <- filter(dat, height > 85)
gt_85
```

`select()` allows you to select columns by name:

```{r}
# columns by name
player_height <- select(dat, player, height)
```


### Your turn:

- use `slice()` to subset the data by selecting the first 5 rows.
- use `slice()` to subset the data by selecting rows 10, 15, 20, ..., 50.
- use `slice()` to subset the data by selecting the last 5 rows.
- use `filter()` to subset those players with height less than 70 inches tall.
- use `filter()` to subset rows of Golden State Warriors ('GSW').
- use `filter()` to subset rows of GSW centers ('C').
- use `filter()` and then `select()`, to subset rows of lakers ('LAL'), and 
then display their names.
- use `filter()` and then `select()`, to display the name and salary, 
of GSW point guards
- find how to select the name, age, and team, of players with more than 
10 years of experience, making 10 million dollars or less.
- find how to select the name, team, height, and weight, of rookie players, 
20 years old, displaying only the first five occurrences (i.e. rows)


-----

## Adding new variables: `mutate()`

Another basic verb is `mutate()` which allows you to add new variables.
Let's create a small data frame for the warriors with three columns: 
`player`, `height`, and `weight`:

```{r}
# creating a small data frame step by step
gsw <- filter(dat, team == 'GSW')
gsw <- select(gsw, player, height, weight)
gsw <- slice(gsw, c(4, 8, 10, 14, 15))
gsw
```

Now, let's use `mutate()` to (temporarily) add a column with the ratio 
`height / weight`:

```{r}
mutate(gsw, height / weight)
```

You can also give a new name, like: `ht_wt = height / weight`:

```{r}
mutate(gsw, ht_wt = height / weight)
```

In order to permanently change the data, you need to assign the changes to 
an object:

```{r}
gsw2 <- mutate(gsw, ht_m = height * 0.0254, wt_kg = weight * 0.4536)
gsw2
```


## Reordering rows: `arrange()`

The next basic verb of `"dplyr"` is `arrange()` which allows you to reorder 
rows. For example, here's how to arrange the rows of `gsw` by `height`

```{r}
# order rows by height (increasingly)
arrange(gsw, height)
```

By default `arrange()` sorts rows in increasing. To arrande rows in 
descending order you need to use the auxiliary function `desc()`.

```{r}
# order rows by height (decreasingly)
arrange(gsw, desc(height))
```

```{r}
# order rows by height, and then weight
arrange(gsw, height, weight)
```

-----

### Your Turn

- using the data frame `gsw`, add a new variable `product` with the product of 
`height` and `weight`.

- create a new data frame `gsw3`, by adding columns `log_height` and `log_weight`
with the log transformations of `height` and `weight`.

- use the original data frame to `filter()` and `arrange()` those players with height less than 71 inches tall, in increasing order.

- display the name, team, and salary, of the top-5 highest paid players

- display the name, team, and salary, for the top-5 highest paid players

- display the name, team, and points3, of the top 10 three-point players

- create a data frame `gsw_mpg` of GSW players, that contains variables for 
player name, experience, and `min_per_game` (minutes per game), sorted by `min_per_game` (in descending order)


-----

## Summarizing values with `summarise()`

The next verb is `summarise()`. Conceptually, this involves applying a function
on one or more columns, in order to summarize values. This is probably easier 
to understand with one example.

Say you are interested in calculating the average salary of all NBA players.
To do this "a la dplyr" you use `summarise()`, or its synonym function 
`summarize()`:

```{r}
# average salary of NBA players
summarise(dat, avg_salary = mean(salary))
```

Calculating an average like this seems a bit _verbose_, especially
when you can directly use `mean()` like this:

```{r}
mean(dat$salary)
```

So let's make things a bit more interessting. What if you want to calculate some 
summary statistics for `salary`: min, median, mean, and max?

```{r}
# some stats for salary (dplyr)
summarise(
  dat, 
  min = min(salary),
  median = median(salary),
  avg = mean(salary),
  max = max(salary)
)
```

Well, this may still look like not much. You can do the same in base R 
(there are actually better ways to do this):

```{r}
# some stats for salary (base R)
c(min = min(dat$salary), 
  median = median(dat$salary),
  median = mean(dat$salary),
  max = max(dat$salary))
```


## Grouped operations

To actually appreciate the power of `summarise()`, we need to introduce the 
other major basic verb in `"dplyr"`: `group_by()`. This is the function that 
allows you to do perform data aggregations, or _grouped operations_.

Let's see the combination of `summarise()` and `group_by()` to calculate the
average salary by team:

```{r}
# average salary, grouped by team
summarise(
  group_by(dat, team),
  avg_salary = mean(salary)
)
```

Here's a similar example with the average salary by position:

```{r}
# average salary, grouped by position
summarise(
  group_by(dat, position),
  avg_salary = mean(salary)
)
```

Here's a more fancy example: average weight and height, by position, displayed
in desceding order by average height:

```{r}
arrange(
  summarise(
    group_by(dat, position),
    avg_height = mean(height),
    avg_weight = mean(weight)),
  desc(avg_height)
)
```


### Your turn:

- use `summarise()` to get the largest height value.

- use `summarise()` to get the standard deviation of `points3`.

- use `summarise()` and `group_by()` to display the median of three-points,
by team.

- display the average triple points by team, in ascending order, of the 
bottom-5 teams (worst 3pointer teams)

- obtain the mean and standard deviation of `age`, for Power Forwards, with 5 
and 10 years (including) years of experience.

-----

## First contact with `ggplot()`

The package `"ggplot2"` is probably the most popular package in R to create
_beautiful_ graphics. In contrast to the functions in the base package 
`"graphcics"`, the package `"ggplot2`" follows a somewhat different philosophy, 
and it tries to be more consistent and modular as possible.

- The main function in `"ggplot2"` is `ggplot()`
- The main input to `ggplot()` is a data frame object.
- You can use the internal function `aes()` to specify what columns of the data 
frame will be used for the graphical elements of the plot.
- You must specify what kind of _geometric objects_ or __geoms__ will be
displayed: e.g. `geom_point()`, `geom_bar()`, `geom_boxpot()`.
- Pretty much anything else that you want to add to your plot is controlled 
by auxiliary functions, especially those things that have to with the format,
rather than the underlying data.
- The construction of a ggplot is done by _adding layers_ with the `+` operator.


### Scatterplots

Let's start with a scatterplot of `salary` and `points`

```{r}
# scatterplot (option 1)
ggplot(data = dat) +
  geom_point(aes(x = points, y = salary))
```

- `ggplot()` creates an object of class `"ggplot"`
- the main input for `ggplot()` is `data` which must be a data frame
- then we use the `"+"` operator to add a layer
- the geometric object (geom) are points: `geom_points()`
- `aes()` is used to specify the `x` and `y` coordinates, by taking columns
`points` and `salary` from the data frame

The same scatterplot can also be created with this alternative, and more 
common use of `ggplot()`

```{r eval = FALSE}
# scatterplot (option 2)
ggplot(data = dat, aes(x = points, y = salary)) +
  geom_point()
```

Say you want to color code the points in terms of the `position`

```{r}
# colored scatterplot 
ggplot(data = dat, aes(x = points, y = salary)) +
  geom_point(aes(color = position))
```

Maybe you wan to modify the size of the dots in terms of `points3`:

```{r}
# sized and colored scatterplot 
ggplot(data = dat, aes(x = points, y = salary)) +
  geom_point(aes(color = position, size = points3))
```

To add some transparency effect to the dots, you can use the `alpha` parameter

```{r}
# sized and colored scatterplot 
ggplot(data = dat, aes(x = points, y = salary)) +
  geom_point(aes(color = position, size = points3), alpha = 0.7)
```

Notice that `alpha` was specified outside `aes()`. This is because we are not 
using any column for the `alpha` transparency values.


### Your turn:

- Use the data frame `gsw` to make a scatterplot of `height` and `weight`
- Find out how to make another scatterplot of `height` and `weight`, 
 using `geom_text()` to display the names of the players
- Get a scatter plot of `height` and `weight`, for ALL the warriors, displaying
their names with `geom_label()`
- Get a density plot of `salary` (for all NBA players)
- Get a histogram of `points2` with binwidth of 50 (for all NBA players)
- Get a barchart of the `position` frequencies (for all NBA players)
- Make a scatterplot of `experience` and `salary` of all centers, and use 
`geom_smooth()` to add a regression line
- Repeat the same scatterplot of `experience` and `salary` of all centers, but 
now use `geom_smooth()` to add a loess line


## Faceting

One of the most attractive features of `"ggplot2"` is the ability to display
multiple __facets__. The idea of facets is to divide a plot into subplots 
based on the values of one or more categorical (or discrete) variables.

Here's an example. What if you want to get scatterplots of `points` and `salary` 
separated (or grouped) by `position`? This is where faceting comes handy,
and you can use `facet_warp()` for this purpose:

```{r}
# scatterplot by position
ggplot(data = dat, aes(x = points, y = salary)) +
  geom_point() +
  facet_wrap(~ position)
```

The other faceting function is `facet_grid()`, which allows you to control the
layout of the facets (by rows, by columns, etc)

```{r}
# scatterplot by position
ggplot(data = dat, aes(x = points, y = salary)) +
  geom_point(aes(color = position), alpha = 0.7) +
  facet_grid(~ position) +
  geom_smooth(method = loess)


# scatterplot by position
ggplot(data = dat, aes(x = points, y = salary)) +
  geom_point(aes(color = position), alpha = 0.7) +
  facet_grid(position ~ .) +
  geom_smooth(method = loess)
```

### Your turn:

- Make scatterplots of `experience` and `salary` faceting by `position`
- Make scatterplots of `experience` and `salary` faceting by `team`
- Make density plots of `age` faceting by `team`
- Make scatterplots of `height` and `weight` faceting by `position`
- Make scatterplots of `height` and `weight`, with a 2-dimensional density, 
`geom_density2d()`, faceting by `position`
