---
title: "Lab 3: Vectors and other data structures"
subtitle: "Stat 133, Fall 2017"
author: "Gaston Sanchez"
output: github_document
urlcolor: blue
---

> ### Learning Objectives
>
> - Work with vectors of different data types
> - Understand the concept of _atomic_ structures
> - Learn how to subset and slice R vectors
> - Understand the concept of _vectorization_
> - Understand _recycling_ rules in R

------

## Getting the Data File

In this lab, you are going to work with a handful of variables about the 
U.S. States:

- _State_: names of the States
- _Capital_: names of the capitals
- _Area_: total area in square kilometers
- _Water_: water area in square kilometers
- _Seats_: number of House seats

The data comes from the wikipedia page of the List of States and
territories of the US: 
[https://en.wikipedia.org/wiki/List_of_states_and_territories_of_the_United_States](https://en.wikipedia.org/wiki/List_of_states_and_territories_of_the_United_States)


Open a new session in Rstudio, and make sure you have a clean workspace:

```r
# remove existing objects
rm(list = ls())
```

Use the following code to read the data contained in the file 
`usa-states.RData` (from the course's github repository)

```r
# download RData file into your working directory
rdata <- "https://github.com/ucb-stat133/stat133-fall-2017/raw/master/data/usa-states.RData"
download.file(url = rdata, destfile = 'usa-states.RData')

# load data in your R session
load('usa-states.RData')

# list the available objects with ls()
ls()
```

### What's happening in the code above?

The function `download.file()` allows you to download any type of file from 
the Web. In this case you are downloading the file called `usa-states.RData` 
which is located in the github repository of the course. This file is a binary 
file. To be more precise, the file extension `.RData` is the default extension 
used by R for its binary native format.

Where does the file is downloaded? The file `usa-states.RData` gets downloaded 
to your __working directory__. If you are curious about what is the current 
directory to which R is paying attention, simply type the function 
`getwd()`---which stands for _get the working directory_.

To load the contents of the binary file into R's session you use `load()`. 
Finally, `ls()` allows you to __list__ all the available R objects.

Check that the following objects are available in your session:

- `state` (names of the States)
- `capital` (names of the capitals)
- `area` (total area in km2)
- `water` (water area in km2)
- `seats` (number of house seats)

__A note on `.RData` files.__ Most of the data sets you are going to be
working with in real life are going to be stored in some sort of text file. 
So you probably won't be handling many R binary files (i.e. `.RData` files). 
However, R binary files are an interesting option for saving intermediate 
results, and/or for saving R objects (as R objects).



### Inspecting the data objects

Once you have some data objects to work with, the first step is to
inspect some of their characteristics. R has various of functions that
allow you to examine objects: 

- `typeof()` type of storage of any object
- `class()` gives you the class of the object
- `str()` displays the structure of an object in a compact way
- `mode()` gives the data type (as used in R)
- `object.size()` gives an estimate of the memory space used by an object
- `length()` gives the length (i.e. number of elements)
- `head()` take a peek at the first elements
- `tail()` take a peek at the last elements
- `summary()` shows a summary of a given object

Your turn. Find out what is the class of each of the objects `state`, `capital`, etc:

```{r}
# your code here
```

Now use `length()`, `head()`, `tail()`, and `summary()` to start exploring
the content of the objects:

```{r}
# your code here
```

Do all the objects have the same length?


## Vectors in R

Vectors are the most basic type of data structures in R. Learning how to
manipulate data structures in R requires you to start learning 
how to manipulate vectors.

How do you know if any of the loaded objects is a vector? Using `class()` does 
not explicitly tell you whether an R object is a vector or not. A more explicit 
way to check if `state` or `capital` are vectors is with the function 
`is.vector()`.

```{r}
# check whether the loaded objects are vectors
```

Remember that R vectors are __atomic structures__, which is just the fancy
name to indicate that all the elements of a vector must be of the same type,
either all numbers, all characters, or all logical values. 

How do you know that a given vector is of a certain data type? One function 
to answer this question is `typeof()`

```r
typeof(state)
typeof(capital)
typeof(area)
typeof(water)
```


## Manipulating Vectors: Subsetting

Subsetting refers to extracting elements of a vector (or another R object). 
To do so, you use what is known as __bracket notation__. This implies using 
(square) brackets `[ ]` to get access to the elements of a vector, for 
instance:

```r
# first element of 'state'
state[1]

# first five elements of 'state'
state[1:5]
```

What type of things can you specify inside the brackets? Basically:

- numeric vectors
- logical vectors (the length of the logical vector must match the length
of the vector to be subset)
- character vectors (if the elements have names)


### Subsetting with Numeric Indices

Here are some subsetting examples using a numeric vector inside the 
brackets:

```r
# fifth element of 'state'
state[5]

# numeric range
state[2:8]

# numeric vector
state[c(1, 3, 5, 7)]

# different order
state[c(20, 9, 10, 50)]

# third element (four times)
state[rep(3, 4)]
```

Try these. What happens if you specify: 

- an index of zero: `state[0]`?
- a negative index: `state[-1]`?
- various negative indices: `state[-c(1,2,3,4)]`?
- an index greater than the length of the vector: `state[53]`?
- repeated indices: `state[c(1,2,2,3,3,3)]`?


### Subsetting with Logical Indices

Logical subsetting involves using a logical vector inside the brackets.
Learning about _logical subsetting_ is a fundamental survival skill. 
This kind of subsetting is __very powerful__ because it allows you to 
extract elements based on some logical condition. 

Here's a toy example of logical subsetting:

```{r}
# dummy vector
a <- c(5, 6, 7, 8)

# logical subsetting
a[c(TRUE, FALSE, TRUE, FALSE)]
```

To do logical subsetting, the vector that you put inside the brackets,
must match the length of the manipulated vector. The elements of the vector
that are subset are those which match the logical value `TRUE`.

```r
# your turn
a[c(TRUE, TRUE, TRUE, TRUE)]
a[c(TRUE, TRUE, FALSE, FALSE)]
a[c(FALSE, FALSE, TRUE, TRUE)]
a[c(TRUE, FALSE, TRUE, FALSE)]
a[c(FALSE, FALSE, FALSE, FALSE)]
```

When subsetting a vector logically, most of the times you won't really be 
providing an explicit vector of `TRUE`'s and `FALSE`s. Just imagine having a 
vector of 100 or 1000 or 1000000 elements, and trying to do logical subsetting 
by manually creating a logical vector of the same length. Instead, you will be 
providing a logical condition or a comparison operation that returns a logical 
vector.

A comparison operation occurs when you use comparison operators such as: 

- `>` greater than
- `>=` greater than or equal
- `<` less than
- `<=` less than or equal
- `==` equal
- `!=` different

```{r}
# elements greater than 6
a > 6

# elements different from 6
a != 6
```

Notice that a comparison operation always returns a logical vector.
Here's how you actually extract the values of a vector based on a logical indexing:

```{r}
a <- c(5, 6, 7, 8)

# elements greater than 6
a[a > 6]

# elements less then 8
a[a < 8]

# elements less than or equal to 6
a[a <= 6]

# elements different from 5
a[a != 5]
```

Logical conditions, in turn, use a logical operator:

- `&` AND
- `|` OR
- `!` negation

```{r}
# AND
TRUE & TRUE
TRUE & FALSE
FALSE & FALSE

# OR
TRUE | TRUE
TRUE | FALSE
FALSE | FALSE

# NOT
!TRUE
!FALSE
```


Here are more examples of logical subsetting with the US States data vectors:

```r
# area of California
area[state == 'California']

# name of states with areas greater than 400,000 square km
state[area > 400000]

# name of states with areas between 100,000 and 125,000 square km
state[area > 100000 & area < 125000]
```


### Your turn

Write commands to answer the following questions:

```{r}
# name of the state with largest area


# name of the state with smallest area


# name of the state with largest number of seats


# capital of the state with the smallest water area


#
```

### Subsetting with Character Vectors

A third type of subsetting involves passing a character vector inside brackets.
When you do this, the characters are supposed to be names of the manipulated
vector.

None of the vectors `state`, `capital`, `area`, `water`, and `seats` have names.
You can confirm that with the `names()` function applied on any of the vectors:

```r
names(state)
```

Create a new vector `total` by adding `area` and `water`, and then assign 
`state` as the names of `total`

```r
# create 'total


# assign 'state' as names of 'total'


```

You should have a vector `total` with named elements. Now you can use
character subsetting:

```r
total["Alabama"]

total[c("California", "Oregon", "Washington")]

total[c("Texas", "Alaska")]
```


### Some plotting

Use the function `plot()` to make a scatterplot of `area` and `water`

```r
plot(area, water)
```

Keep in mind tht `plot()` is a generic function. This means that the behavior
of `plot()` depends on the type of input. When you pass two numeric vectors
to `plot()` it will create a scatter plot.

Looking at the generated plot, can you see any issues?

To get a better display of the scatterplot, let's create two vectors
`log_area` and `log_water` by transforming `area` and `water` with the 
logarithm function `log()`

```r
log_area <- log(area)
log_water <- log(water)
```

Make another scatterplot but now use the log-transformed vectors:

```{r, eval = FALSE}
plot(log_area, log_water)
```

To add the names of the states in the plot, you can use `text()`:

```{r, eval = FALSE}
plot(log_area, log_water)
text(log_area, log_water, labels = state)
```

Now we have another problem. The labels in the plot are a bit messy. A quick
and dirty fix is to use `abbreviate()` to shorten the displayed names:

```{r, eval = FALSE}
plot(log_area, log_water)
text(log_area, log_water, labels = abbreviate(state))
```


## Vectorization

When you create the vectors `log_area <- log(area)` and 
`log_water <- log(water)`, what you're doing is applying a function to a 
vector, which in turn acts on all elements of the vector.

This is called __Vectorization__ in R parlance. Most functions that operate 
with vectors in R are __vectorized__ functions. This means that an action 
is applied to all elements of the vector without the need to explicitly type 
commands to traverse all the elements.

In many other programming languages, you would have to use a set of commands 
to loop over each element of a vector (or list of numbers) to transform them. 
But not in R. 

Another example of vectorization would be the calculation of the square root 
of all the elements in `area` and `water`:

```r
sqrt(area)
sqrt(water)
```

Or the sum of `area` plus `water` to get the total area:
```r
area + water
```


### Why should you care about vectorization?

If you are new to programming, learning about R's vectorization will be very
natural (you won't stop to think about it too much). If you have some previous
programming experience in other languages (e.g. C, python, perl), you know 
that vectorization does not tend to be a native thing.

Vectorization is essential in R. It saves you from typing many lines of code,
and you will exploit vectorization with other useful functions known as the 
_apply_ family functions (we'll talk about them later in the course).



## Recycling

Closely related with the concept of _vectorization_ we have the notion of
__Recycling__. To explain _recycling_ let's see an example.

`area` and `water` are given in square kilometers, but what if you need to
obtain the areas in square miles?. Let's create two new vectors 
`area_square_miles` and `water_square_miles` with the converted values in 
square miles. To convert from square kilometers to aquare miles use the
following conversion:
1 sqr km = 0.386 sqr mi

```{r}
# your code here
```

What you just did (assuming that you did things correctly) is called 
__Recycling__. To understand this concept, you need to remember that R does 
not have a data structure for scalars (single numbers). Scalars are in reality
vectors of length 1.

Converting square kms to square miles requires this operation: `area * 0.836`.
Although it may not be obvious, we are multiplying two vectors: `area` and 
`0.836`. Moreover (and more important) __we are multiplying two vectors of 
different lengths!__. So how does R know what to do in this cases?

Well, R uses the __recycling rule__, which takes the shorter vector (in this
case `0.386`) and recycles its elements to form a temporary vector that matches
the length of the longer vector (i.e. `area`).


### Another recycling example

Here's another example of recycling. Areas of elements in an odd number 
position will be transformed to square miles; areas of elements in an even 
number position will be transformed to acres:

```r
units <- c(0.386, 247.105)
new_area <- area * units
```

The elements of `units` are recycled and repeated as many times as elements
in `area`. The previous command is equivalent to this:

```r
new_units <- rep(c(0.386, 247.105), length.out = length(area))
area * new_units
```

### Your turn

Make a scatterplot of `area_square_miles` and `water_square_miles`
transforming the vectors into log scale (use `log()`)

```{r}
# your scatter plot
```



-----

## Factors

As mentioned before, vectors are the most essential type of data structure
in R. They are _atomic_ structures (can contain only one type of data):
integers, real numbers, logical values, characters, complex numbers.

Related to vectors, there is another important data structure in R called
__factor__. Factors are data structures exclusively designed to handle 
categorical data.

The term _factor_ as used in R for handling categorical variables, comes from 
the terminology used in _Analysis of Variance_, commonly referred to as ANOVA. 
In this statistical method, a categorical variable is commonly referred to as _factor_ and its categories are known as _levels_.


### Creating Factors

To create a factor you use the homonym function `factor()`, which takes a 
vector as input. The vector can be either numeric, character or logical. 

```{r}
# numeric vector
num_vector <- c(1, 2, 3, 1, 2, 3, 2)

# creating a factor from num_vector
first_factor <- factor(num_vector)

first_factor
```

You can also obtain a factor from a character vector:

```{r}
# string vector
str_vector <- c('a', 'b', 'c', 'b', 'c', 'a', 'c', 'b')

str_vector

# creating a factor from str_vector
second_factor <- factor(str_vector)

second_factor
```

Notice how `str_vector` and `second_factor` are displayed. Even though the 
elements are the same in both the vector and the factor, they are printed in 
different formats. The letters in the string vector are displayed with quotes, 
while the letters in the factor are printed without quotes.


### How does R store factors?

Under the hood, a factor is internally stored using two arrays: one is an 
integer array containing the values of categories, the other array is the 
"levels" which has the names of categories which are mapped to the integers.

One way to confirm that the values of the categories are mapped as integers 
is by using the function `storage.mode()`

```{r}
# storage of factor
storage.mode(first_factor)
```

### Manipulating Factors

Because factors are internally stored as integers, you can manipulate factors
as any other vector:

```{r}
first_factor[1:5]

first_factor[c(1, 3, 5)]

first_factor[rep(1, 5)]

second_factor[second_factor == 'a']

second_factor[second_factor == 'b']
```

-----
