---
title: "Lab 7: Simple Functions and Conditionals"
subtitle: "Stat 133, Fall 2017"
author: "Gaston Sanchez"
output: github_document
fontsize: 12pt
urlcolor: blue
---

> ### Learning Objectives
>
> - Learn to how to write simple functions
> - Get into the habit of writing simple functions
> - Making sure your functions work
> - Use conditionals if-then-else

------


## Simple Math Functions

Consider the following mathematical functions:

- $f(x) = x^2$
- $g(x) = 2x + 5$

Write two functions `f()` and `g()` based on the previous equations.

```{r}
# your function f()

```

```{r}
# your function g()

```

\bigskip

Test your functions with:

```r
f(2)
f(-5)

g(0)
g(-5/2)
```

Write code to create the following composite functions:

- `fog()` for the composite function: $f \circ g(x)$
- `gof()` for the composite function: $g \circ f(x)$

```{r}
# your function fog()

```

```{r}
# your function gof()

```

Test your composite functions with:

```r
fog(2)
fog(-5)

gof(0)
gof(-5/2)
```


-----


## Pythagoras

The pythagoras formula is used to compute the length of the hypotenuse, $c$, of 
a right triangle with legs of length $a$ and $b$.

![hypotenuse](https://wikimedia.org/api/rest_v1/media/math/render/svg/5fd521cee81d583ce94bf6710984cc2a9eb7c3da)

Write a function `pythagoras()` that takes two arguments `a` and `b`, and 
returns the length of the hypotenuse:

```{r}
# your pythagoras() function

```

Test your `pythagoras()` with two leg values: `pythagoras(3, 4)`

```{r}
# your test

```

Modify your function `pythagoras()` so that argument `b` takes the same value of argument `a`. Test it with just one leg value: `pythagoras(5)`

```{r}
# your test

```


-----


## Unit Conversion Formulas

The following exercises involve writing simple functions to convert from one 
type of unit to other.


### Miles to Kilometers

Write a function `miles2kms()` that converts miles into kilometers: 1 mile
is equal to 1.6 kilometers. Give the argument a default value of 1.

```{r}
miles2kms <- function() {
  # fill in
}
```

Use `miles2kms()` to obtain mile conversions, in order to create a table 
(i.e. data frame) like the one below. The first ten rows range from 1 to 10 miles, 
and then from 10 to 100 in 10 mile steps. The second column corresponds to kms.

| miles | kms |
|-------|-----|
|   1   | 1.6 |
|   2   | 3.2 |
|  ...  | ... |
|  10   | 16  |
|  20   | 32  |
|  ...  | ... |
| 100   | 160 |


### Gallons to Liters, and viceversa

Write a function `gallon2liters()` that converts gallons to liters: 1 gallon
is equal to 3.78541 liters:

```{r}
# your gallons2liters() function

```

Use `gallon2liters()` to make an inverse function `liters2gallons()`.

```{r}
# your liters2gallons() function

```

Use `liters2gallons()` to obtain liter conversions, in order to create a table 
(i.e. data frame) like the one below. The first ten rows range from 1 to 10 liters, 
and then from 10 to 100 in 10 liter steps. The second column corresponds to gallons.

| liters | gallons   |
|--------|-----------|
|   1    | 0.2641722 |
|   2    | 0.5283444 |
|  ...   | ...       |
|  10    | 2.6417218 |
|  20    | 5.283444  |
|  ...   | ...       |
| 100    | 26.417218 |



### Seconds to Years

According to Wikipedia, in 2015 the life expectancy of a person born in the US 
was 79 years. Consider the following question: Can a newborn baby in USA expect 
to live for one billion ($10^9$) seconds? 

To answer this question, write a function `seconds2years()` that takes
a number in seconds and returns the equivalent number of years.
Test the function with `seconds2years(1000000000)`

```{r}
# your code

```


-----


## Gaussian Function

The Gaussian (Normal) function, given in the equation below, is one of the 
most widely used functions in science and statistics:

![gaussian function](https://wikimedia.org/api/rest_v1/media/math/render/svg/8aa9ff808602c27f1d9d63d7b2c115388a34f190)

The parameters $\sigma$ and $\mu$ are real numbers, where $\sigma$ must be greater than 
zero. 

Make a function `gaussian()` that takes three arguments: `x`, `m`, and `s`. 
Evaluate the function with $m = 0$, $s = 2$, and $x = 1$.

```{r}
# your code

```

Test your `gaussian()` function and compare it with the R function `dnorm()`

```{r}
# compare with dnorm()
dnorm(x = 1, mean = 0, sd = 2)
```

Now try `gaussian()` with a vector `seq(-4.5, 4.5, by = 0.1)`, and pass the values
to `plot()` to get a normal curve. Here's code with values obtained from `dnorm()`

```{r}
# you should get a plot like this one
x_values <- seq(from = -4.5, to = 4.5, by = 0.1)
y_values <- dnorm(x_values, mean = 0, sd = 2)
plot(x_values, y_values, las = 1, type = "l", lwd = 2)
```

Your turn:

```{r}
# your code
```


-----


### Polynomials

In this problem we want to see whether the graph of a given polynomial
will cross or touch the _x_ axis in a given interval. 

Let's begin with the polynomial: $f(x) = x^2 (x - 1)$. The first thing 
to do is write a function for the polynomial, for instance:

```{r}
poly1 <- function(x) {
  (x^2) * (x - 1)
}
```

Once you have a function for the polynomial, you can create a set of pairs of
points $x$ and $y = f(x)$, and then use them to graph the polynomial

```{r}
# set of points
x <- seq(-4, 4, length.out = 20)
y <- poly1(x)

# graph polynomial
plot(x, y, type = 'l', lwd = 3, col = "#FB7215", las = 1)
abline(h = 0, v = 0, col = '#888888aa', lwd = 1.5)
title(main = expression(paste(f(x), ' = ', x^2, (x - 1))))
```

Write functions and graph the following polynomials in the x-axis interval
-4 to 4:

a. $f(x) = x^3$ 
b. $f(x) = (x^2 - 1)(x + 3)^3$
c. $f(x) = (x^2 - 1)(x^2 - 9)$


-----


## Descriptive Statistics

Write a function `descriptive()` that takes a numeric vector as input, and 
returns a named vector with the following descriptive statistics:

- `min`: minimum
- `q1`: first quartile (Q2)
- `median`: median
- `mean`: mean
- `q3`: third quartile (Q3)
- `max`: maximum
- `range`: range or span (max - min)
- `iqr`: interquartile range (IQR)
- `sd`: standard deviation

```{r}
# your descriptive() function
descriptive <- function() {
  # fill in

}
```


## Binomial Formula

In [lab01](lab01-R-basics.md) you worked with the formula of the binomial 
probability:

![binomial probability](https://wikimedia.org/api/rest_v1/media/math/render/svg/38d86cba65d40f015a2b807d2b736250805abe45)

where:

- $n$ is the number of (fixed) trials
- $p$ is the probability of success on each trial
- $1 - p$ is the probability of failure on each trial
- $k$ is a variable that represents the number of successes out of $n$ trials
- the first term in parenthesis is not a fraction, it is the number of combinations in which $k$ success can occur in $n$ trials

R provides the `choose()` function to compute the number of combinations: 

![combinations](https://wikimedia.org/api/rest_v1/media/math/render/svg/08bdf0fff474c26293414f9eb01ab4bc73ef941f)

For instance, the number of combinations in which $k$ = 2 
success can occur in $n$ = 5 trials is:
```{r}
choose(n = 5, k = 2)
```

Combinations are typically expressed in terms of factorials as:

![combs](https://wikimedia.org/api/rest_v1/media/math/render/svg/813f7124a61dac205542db3f8491b36cb306453a)

Conveniently, R also provides the function `factorial()` to calculate the factorial of an integer:

```{r}
factorial(4)
```


## Your turn: write a binomial function

- Use `factorial()` to create a `combinations()` function. Your function should have arguments `n` and `k`:
```r
# you should be able to call combinations() like this
combinations(n = 5, k = 2)
```

- Use your `combinations()` function to write a `binom_prob()` function that 
computes binomial probabilities. The arguments should be `n`, `k`, 
and `prob`. Here's an example of you should be able to call the function:
```r
# probability of getting 2 successes in 5 trials
# (assuming prob of success = 0.5)
binom_prob(n = 5, k = 2, prob = 0.5)
```

- Use `binom_prob()` to obtain the probability of getting more than 3 heads in 5 tosses
with a biased coin of 35% chance of heads.

- Consider rolling a fair die 10 times. Use `binom_prob()` to find the probability 
of getting exactly 3 sixes?


------


## If Conditionals

Write R code that will "squish" a number into the interval [0, 100], so that a 
number less than 0 is replaced by 0 and a number greater than 100 is replaced 
by 100.

```{r}
z <- 100*pi
# Fill in the following if-else statements. You may (or may not) 
# have to add or subtract else if or else statements.
if (TRUE) { # Replace TRUE with a condition.
  
} else if (TRUE) { # Replace TRUE with a condition.
  
} else {
  
}
```

### Multiple If's

A common situation involves working with multiple conditions at the same time. 
You can chain multiple if-else statements:

```{r}
y <- 1 # Change this value!

if (y > 0) {
  print("positive")
} else if (y < 0) {
  print("negative")
} else {
  print("zero?")
}
```


## Even number

Write a function `is_even()` that determines whether a number is even 
(i.e. multiple of 2). If the input number is even, the output should be 
`TRUE`. If the input number is odd, the output should be `FALSE`. 
If the input is not a number, the output should be `NA` 

Test your function:
```{r eval = FALSE}
# even number
is_even(10)

# odd number
is_even(33)

# not a number
is_even('a')
```


## Odd number

Use your function `is_even()` to write a function `is_odd()` that determines 
if a number is odd (i.e. not a multiple of 2). If a number is odd, the output 
should be `TRUE`; if a number is even the output should be `FALSE`; if the 
input is not a number the output should be `NA`

Test `is_odd()` with the following cases:
```{r eval = FALSE}
# odd number
is_odd(1)

# even number
is_odd(4)

# not a number
is_odd('a')
```


## Switch

Working with multiple chained if's becomes cumbersome. Consider the following
example that uses several if's to convert a day of the week into a number:
```{r}
# Convert the day of the week into a number.
day <- "Tuesday" # Change this value!

if (day == 'Sunday') {
  num_day <- 1
} else {
  if (day == "Monday") {
    num_day <- 2
  } else {
    if (day == "Tuesday") {
      num_day <- 3
    } else {
      if (day == "Wednesday") {
        num_day <- 4
      } else {
        if (day == "Thursday") {
          num_day <- 5
        } else {
          if (day == "Friday") {
            num_day <- 6
          } else {
            if (day == "Saturday") {
              num_day <- 7
            }
          }
        }
      }
    }
  }
}


num_day
```

Working with several nested if's like in the example above can be a nigthmare.

In R, you can get rid of many of the braces like this:

```{r}
# Convert the day of the week into a number.
day <- "Tuesday" # Change this value!

if (day == 'Sunday') {
  num_day <- 1
} else if (day == "Monday") {
  num_day <- 2
} else if (day == "Tuesday") {
  num_day <- 3
} else if (day == "Wednesday") {
  num_day <- 4
} else if (day == "Thursday") {
  num_day <- 5
} else if (day == "Friday") {
  num_day <- 6
} else if (day == "Saturday") {
  num_day <- 7
}

num_day
```


### Your turn: a grading function

Write a function `grade()` that takes a `score` argument (i.e. a numeric value between 0 and 100), and returns a letter grade (i.e. character) based on the following grading scheme:

| score      | grade   |
| ---------- | ------- |
| [90 - 100] |  `"A"`  |
| [80 - 90)  |  `"B"`  |
| [70 - 80)  |  `"C"`  |
| [60 - 70)  |  `"D"`  |
| < 60       |  `"F"`  |

You should be able to call your `grade()` function like this:

```{r eval = FALSE}
# grade "A"
grade(score = 90)

# grade "B"
grade(score = 89.9999)

# grade "C"
grade(score = 70.0000001)

# grade "F"
grade(score = 50)
```


Modify your `grade()` function to include a `stop()` condition when the input `score` value is less than zero or greater than 100. The error message should say something like: 

```
"score must be a number between 0 and 100"
```

-----


## The `switch()` function

Consider again the chain of if-then-else's used to convert a name day into a number dat. We have too many if's, and there's a lot of repetition in the code.
If you find yourself using many if-else statements with identical structure for 
slightly different cases, you may want to consider a __switch__ statement 
instead:

```{r}
# Convert the day of the week into a number.
day <- "Tuesday" # Change this value!

switch(day, # The expression to be evaluated.
  Sunday = 1,
  Monday = 2,
  Tuesday = 3,
  Wednesday = 4,
  Thursday = 5,
  Friday = 6,
  Saturday = 7,
  NA) # an (optional) default value if there are no matches
```

Switch statements can also accept integer arguments, which will act as indices 
to choose a corresponding element:

```{r}
# Convert a number into a day of the week.
day_num <- 3 # Change this value!

switch(day_num,
  "Sunday",
  "Monday",
  "Tuesday",
  "Wednesday",
  "Thursday",
  "Friday",
  "Saturday")
```

## Converting Miles to other units

The table below shows the different formulas for converting miles (mi) into other scales:

| Units   | Formula         |
| ------- | --------------- |
| Inches  | mi x 63360      |
| Feet    | mi x 5280       |
| Yards   | mi x 1760       |
| Meters  | mi / 0.00062137 |
| Kms     | mi / 0.62137    |

Write the following five functions for each type of conversion. Each function 
must take one argument `x` with default value: `x = 1`.

- `miles2inches()`
- `miles2feet()`
- `miles2yards()`
- `miles2meters()`
- `miles2kms()`

For example: 
```{r eval = FALSE}
miles2inches(2)
miles2feet(2)
miles2yards(2)
miles2meters(2)
miles2kms(2)
```

-----

## Using switch()

Create a function `convert()` that converts miles into the specified units. 
Use `switch()` and the previously defined functions---`miles2inches()`, 
`miles2feet()`, ..., `miles2kms`---to define `convert()`. Use two arguments: 
`x` and `to`, like this:
```{r, eval=FALSE}
convert(40, to = "in")
```

By default, `to = "km"`, but it can take values such as `"in"`, `"ft"`, `"yd"`, or `"m"`.

Test `convert()` with:
```{r eval=FALSE}
convert(3, "in")
convert(3, "ft")
convert(3, "yd")
convert(3, "m")
convert(3, "km")
```


