---
title: "Lab 6: More dplyr, ggplot2, and files' stuff"
subtitle: "Stat 133, Fall 2017"
author: "Gaston Sanchez"
output: github_document
urlcolor: blue
---

> ### Learning Objectives:
>
> - Work with a more complex file structure
> - Practice exporting tables
> - Practice exporting R output (as is)
> - Practice exporting plot images
> - Learn about `"dplyr"` pipelines
> - Get to know the pipe operator `%>%`
> - Chain dplyr operations with the piper

------

```{r read, echo = FALSE, message = FALSE}
# with "base" read.csv()
dat <- read.csv('../data/nba2017-players.csv', stringsAsFactors = FALSE)
library(dplyr)
```


## Manipulating and Visualizing Data Frames

In this lab, you will continue manipulating data frames with `"dplyr"`, 
and plotting graphics with `"ggplot2"`. In addition, you will also use various
functions to export (or save) tables, images, and R output to external files.

While you follow this lab, you may want to open these cheat sheets:

- [dplyr cheatsheet](../cheat-sheets/data-transformation-cheatsheet.pdf)
- [ggplot2 cheatsheet](../cheat-sheets/ggplot2-cheatsheet-2.1.pdf)


### Filestructure

To help you better prepare for HW03, we want you to practice working with a 
more sophisticated file structure. Follow the steps listed below to create the 
necessary subdirectories like those depicted in this scheme:

```
    lab06/
      README.md
      data/
      code/
      output/
      images/
```

- Open a Bash terminal (e.g. command line or GitBash)
- Change your working directory to a location where you will store all the 
materials for this lab
- Create a directory `lab06` for the lab materials
```bash
mkdir lab06
```
- Change directory to `lab06`
```bash
cd lab06
```
- Make other subdirectories: `data`, `code`, `output`, `images`
```bash
mkdir data code output images
```
- List the contents of `lab06` to confirm that you have all the subdirectories
```bash
ls
```
- Initialize (or create) an empty `README.md` text file
```bash
touch README.md
```
- Change directory to the `data/` folder
```bash
cd data
```
- Download the data file with the command `curl`, and the `-O` option (letter O)
```bash
curl -O https://raw.githubusercontent.com/ucb-stat133/stat133-fall-2017/master/data/nba2017-players.csv
```
- Use `ls` to confirm that the csv file is in `data/`
- Use _word count_ `wc` to count the lines of the csv file
```bash
wc nba2017-players.csv
```
- Take a peek at the first rows of the csv file with `head`
```bash
head nba2017-players.csv
```
- Take a peek at the last 5 rows of the csv file with `tail`
```bash
tail -n 5 nba2017-players.csv
```

-----

# R script file

- Once you have the filestructure for this lab, go to RStudio and open a new 
`R` script file (do NOT confuse with an `Rmd` file).
- Save the `R` script file as `lab06-script.R` in the `code/` folder of `lab06/`

R script files are used to write R code only, using R syntax. In other words,
you should NOT use Markdown syntax or LaTeX inside an R script file. Why?
Because if you run the entire script, R will try to execute all the 
commands, and won't be able to recognize Markdown, LaTeX, yaml, or other 
syntaxes.


## File Header

Let's start with some good coding practices by adding a header to the `R` script file in 
the form of R comments. In general, the header section should contain 
a title, a description of what the script is about, what are the inputs, and 
what are the main outputs produced when executing the code in the script. 
Optionally, you can also include the name of the author, the date, and other details. Something like this:

```
# Title: Short title (one sentence)
# Description: what the script is about (one paragraph or two)
# Input(s): what are the main inputs (list of inputs)
# Output(s): what are the main outputs (list of outputs)
# Author: First Last
# Date: mm-dd-yyyy
```

Think of the header of a script file as the yaml header used in `Rmd` files. The header 
should be the very first thing that appears at the top of the script file.
Personally, I like to surround the header in my R script files with some delimiting
characters that help the reader to visually identify main parts of the script. Here's 
a hypothetical example of a header:

```r
# ===================================================================
# Title: Cleaning Data
# Description:
#   This script performs cleaning tasks and transformations on 
#   various columns of the raw data file.
# Input(s): data file 'raw-data.csv'
# Output(s): data file 'clean-data.csv'
# Author: Gaston Sanchez
# Date: 10-05-2017
# ===================================================================
```

Another good coding practice is to avoid writing very long lines of code.
Most coding style guides stick to a maximum line width of 80 characters, and
this is the [magic number](https://softwareengineering.stackexchange.com/questions/148677/why-is-80-characters-the-standard-limit-for-code-width) that I also use for my scripts.

__Your turn__: Include a header in your R script file, respecting the width 
limit of 80 characters.


## Required Packages

The next thing that you need to include in your script file are the required
packages. Not all script files need packages, but many do. When this is the 
case, loading the packages should be the first lines of code to be executed. 

Include the commands to load the following packages in your script:

```r
# packages
library(readr)    # importing data
library(dplyr)    # data wrangling
library(ggplot2)  # graphics
```

In addition to loading the packages, sometimes you will also need to load 
code from other script files. We won't do that today, but you should know
that this is very common as the complexity and size of your projects grow.


## Exporting some data tables

After the header, and the loading-packages sections, the next part in your 
lab script involves importing the data. In addition to importing a data table,
you are also going to practice exporting tables. That is, writing data tables
to external files.

- Use `read_csv()` from the package `"readr"` to import the data `nba2017-players.csv` in R.
- Create one data frame `warriors` by selecting rows---e.g. `filter()`---of 
Golden State Warriors, arranging rows by salary (increasingly).
- Use the function `write.csv()` to export (or save) the data frame `warriors`
to a data file `warriors.csv` in the `folder/` directory. You will need to 
use a relative path to specify the `file` argument. Also, see how to use 
the argument `row.names` to avoid including a first column of numbers.
- Create another data frame `lakers` by selecting rows of Los Angeles Lakers, this time arranging rows by experience (decreasingly).
- Now use the function `write_csv()` to export (or save) the data frame 
`lakers` to a data file `lakers.csv` in the `folder/` directory. You will also 
need to use a relative path to specify the `file` argument.
- Inspect the contents of the `data/` folder and confirm that the csv files
are there.


## Exporting some R output

After exporting the tables to the corresponding csv files, you will produce 
some summary statistics, and then save the generated output to external 
text files. To do this, you will have to learn about the `sink()` function, 
which sends R output to a specified file.

Say you are interested in exporting the summary statistics of `height` and 
`weight`, exactly in the same way they are displayed by R:

```{r}
summary(dat[ ,c('height', 'weight')])
```

One naive option would be to manually copy the text displayed on the console,
and then paste it to a text file. While this may work, it is labor intensive,
error prone, and highly irreproducible. A better way to achieve this task is 
with the `sink()` function. Here's how:

```r
# divert output to the specified file
sink(file = 'summary-height-weight.txt')
summary(dat[ ,c('height', 'weight')])
sink()
```

The fist call to `sink()` opens a connection to the specified file, and then 
all outputs are diverted to that location. The second call to `sink()`, i.e. 
the one without any arguments, closes the connection.

While you are `sink()`ing output to a specified file, all the results
will be send to such file. In other words, nothing will be printed on the 
console. Only adter the sinking process has finished and the connection is
closed, you will be able to execute commands and see results displayed on R's
console.


### Why sinking?

Why would you ever want to `sink()` R outputs to a file? Why not simply 
display them as part of your `Rmd` file? One good reason for diverting output
to an external file is for convenience. In practice, the reports and documents 
(e.g. papers, executive summaries, slides) of a data analysis project won't
contain everything that you tried, explored, analyzed, and graphed. There 
will be many intermediate results that while relevant for a specific stage
of the analysis cycle, they are innecessary for the final report. So a good way
to keep these intermediate outputs is by exporting them with `sink()`.


__Your turn:__

- Export the output `str()` on the data frame with all the players to a 
text file `data-structure.txt`, in the `output/` folder.
- Export the `summary()` of the entire data frame `warriors` to a text file
`summary-warriors.txt`, in the `output/` folder.
- Export another `summary()` of the entire data frame `lakers` to a text file
`summary-lakers.txt`, in the `output/` folder.


## Exporting some "base" graphs

In the same way that R output as it appears on the console can be exported 
to some files, you can do the same with graphics and plots. Actually, 
saving plot images is much more common than `sink()`ing output.

Base R provides a wide array of functions to save images in most common 
formats:

- `png()`
- `jpeg()`
- `tiff()`
- `bmp()`
- `svg()`
- `pdf()`

Like the writing table functions such as `write.table()` or `write.csv()`, 
and the `sink()` function, the graphics device functions require a file name
to be provided.  Here's how to save a simple scatterpot of `height` and 
`weight` in png format:

```r
# saving a scatterplot in png format
png(filename = "scatterplot-height-weight.png")
plot(dat$height, dat$weight, pch = 20, 
     xlab = 'Height', ylab = 'Height')
dev.off()
```

- The function `png()` tells R to save the image in PNG format, using the 
provided filename.
- Invoking `png()` will open a graphics device; not the graphics device of 
RStudio, so you won't be able to see the graphic.
- The `plot()` function produces the scatterplot.
- The function `dev.off()` closes the graphics device.

__Your turn__:

- Open the help documentation of `png()` and related graphic devices.
- Use `png()` to save a scatterplot of `height` and `weight` with `plot()`.
Save the graph in the `images/` folder.
- Save another version of the scatterplot between `height` and `weight`, but
now try to get an image with higher resolution. Save the plot in `images/`.
- Save a histogram in JPEG format of `age` with dimensions (width x height) 
600 x 400 pixels. Save the plot in `images/`.
- Use `pdf()` to save the previous histogram of `age` in PDF format,
with dimensions (width x height) 7 x 5 inches.


## Exporting some ggplots

The package `"ggplot2"` comes with a wrapper function `ggsave()` that allows
you to save ggplot graphics to a specified file. By default, `ggsave()` saves
images in PDF format.

- Use `ggplot()` to make a scatterplot of `points` and `salary`,
and store it in a ggplot object named `gg_pts_salary`.
Then use `ggsave()` to save the plot with dimensions (width x height) 
7 x 5 inches; in the `images/` folder as `points_salary.pdf`

- Use `ggplot()` to create a scatterplot of `height` and `weight`, 
faceting by `position`. Store this in a ggplot object `gg_ht_wt_positions`
Then use `ggsave()` to save the plot with dimensions (width x height) 
6 x 4 inches; in the `images/` folder as `height_weight_by_position.pdf`


-----

# More `"dplyr"`

The last part of this lab involves working the pipe operator `%>%` which 
allows you write function calls in a more human-readable way. This becomes
extremely useful in `"dplyr"` operations that require many steps.

The behavior of `"dplyr"` is functional in the sense that function calls don't 
have side-effects. You must always save their results in order to keep them 
in an object (in memory). This doesn't lead to particularly elegant code, 
especially if you want to do many operations at once. 
You either have to do it step-by-step:

```{r}
# manipulation step-by-step
dat1 <- group_by(dat, team)
dat2 <- select(dat1, team, height, weight)
dat3 <- summarise(dat2,
  avg_height = mean(height, na.rm = TRUE),
  avg_weight = mean(weight, na.rm = TRUE))
dat4 <- arrange(dat3, avg_height)
dat4
```

Or if you don't want to name the intermediate results, you need to wrap the 
function calls inside each other:

```{r}
# inside-out style (hard to read)
arrange(
  summarise(
    select(
      group_by(dat, team),
      team, height, weight
    ),
    avg_height = mean(height, na.rm = TRUE),
    avg_weight = mean(weight, na.rm = TRUE)    
  ),
  avg_height
)
```

This is difficult to read because the order of the operations is from inside 
to out. Thus, the arguments are a long way away from the function. 
To get around this problem, `"dplyr"` provides the `%>%` operator from 
`"magrittr"`.

`x %>% f(y)` turns into `f(x, y)` so you can use it to rewrite multiple 
operations that you can read left-to-right, top-to-bottom:

```{r}
# using %>%
dat %>% 
  group_by(team) %>%
  select(team, height, weight) %>%
  summarise(
    avg_height = mean(height, na.rm = TRUE),
    avg_weight = mean(weight, na.rm = TRUE)) %>%
  arrange(avg_height)
```


Use the piper operator `"%>%"` to perform the following operations:

- display the player names of Lakers `'LAL'`

- display the name and salary of GSW point guards 'PG'

- dislay the name, age, and team, of players with more than 10 years 
of experience, making 10 million dollars or less.

- select the name, team, height, and weight, of rookie players, 
20 years old, displaying only the first five occurrences (i.e. rows)

- create a data frame `gsw_mpg` of GSW players, that contains variables for 
`player` name, `experience`, and `min_per_game` (minutes per game), 
sorted by `min_per_game` (in descending order)

- display the average triple points by team, in ascending order, of the 
bottom-5 teams (worst 3pointer teams)

- obtain the mean and standard deviation of `age`, for Power Forwards, with 5 
and 10 years (including) years of experience
