---
title: "Lab 5: Manipulating data frames with dplyr"
subtitle: "Stat 133, Fall 2017"
author: "Gaston Sanchez"
output: github_document
urlcolor: blue
---

> ### Learning Objectives:
>
> - Get started with `dplyr`
> - Get to know the basic dplyr verbs:
> - `slice()`, `filter()`, `select()`
> - `mutate()`
> - `arrange()`
> - `summarise()`
> - `group_by()`

------

## Manipulating Data Frames

Last week you started to manipulate data tables 
(under the class of `"data.frame"` objects) using bracket notation, 
`dat[ , ]`, and the dollar operator, `dat$name`, in order to select specific 
rows, columns, or cells.

In this lab, you will start learning about another approach to manipulate tables
in R using the functionality of the package `"dplyr"`. This is a fairly recent
paradigm introduced a couple of years ago, but it is based on more than a decade 
of research and work lead by Hadley Wickham.


## NBA Players Data

```{r nba, echo = FALSE}
dat <- read.csv('../data/nba2017-players.csv', stringsAsFactors = FALSE)
```

The data file for this lab is the same you used last week: `nba2017-players.csv`, 
whic is located in the `data/` folder of the course github repository. 
I assume that you already downloaded a copy of the csv file to your computer. 
If that is not the case, here's one option to get your own copy:

```r
# download RData file into your working directory
github <- "https://github.com/ucb-stat133/stat133-fall-2017/raw/master/"
csv <- "data/nba2017-players.csv"
download.file(url = paste0(github, csv), destfile = 'nba2017-players.csv')
```

To import the data in R you can use the base function `read.csv()`, or you 
can also use `read_csv()` from the package `"readr"`:

```{r read, eval = FALSE}
# with "base" read.csv()
dat <- read.csv('nba2017-players.csv', stringsAsFactors = FALSE)

# with "readr" read_csv()
dat <- read_csv('nba2017-players.csv')
```


-----


## Introduction to `"dplyr"`

In order to use the functions in `"dplyr"`, you must load the package first.
Here I'm assuming that you already installed the package. If that's not the 
case then run on the console the command below (do NOT include this in your `Rmd`):

```r
install.packages("dplyr")
```

Remember that you only need to install a package once! After a package has been 
installed in your machine, there is no need to call `install.packages()` again on 
the same package. What you should always invoke in order to use the functions 
in a package is the `library()` function:

```{r warning=FALSE, message=FALSE}
library(dplyr)
```

__About loading packages:__ Another rule to keep in mind is to always load any 
required packages at the very top of your script files (`.R` or `.Rmd` or `.Rnw` files). 
Avoid calling the `library()` function in the middle of a script. Instead, 
load all the packages before anything else.


### Basic `"dplyr"` verbs

To make the learning process of `"dplyr"` gentler, Hadley Wickham proposes 
beginning with a set of five _basic verbs_ or operations for data frames 
(each verb corresponds to a function in `"dplyr"`):

- __filter__: keep rows matching criteria
- __select__: pick columns by name
- __arrange__: reorder rows
- __mutate__: add new variables
- __summarise__: reduce variables to values

-----

## Filtering, slicing, and selecting

I've modified Hadley's list of verbs a little bit. The first _verbs_ I want you 
to learn about `"dplyr"` are `filter()`, `slice()`, and 
`select()`. Why? Because these are the functions to select a subset of rows 
and columns, which are the most essential type of operations when wrangling data 
frames.

`slice()` allows you to select rows by position:

```{r}
# first three rows
slice(dat, 1:3)
```

`filter()` allows you to select rows by condition:

```{r}
# subset rows given a condition
filter(dat, height > 85)
```

`select()` allows you to select columns by name:

```{r}
# columns by name
player_height <- select(dat, player, height)
```


### Your turn:

- use `slice()` to subset the data by selecting the first 5 rows.
- use `slice()` to subset the data by selecting rows 10, 15, 20, ..., 50.
- use `slice()` to subset the data by selecting the last 5 rows.
- use `filter()` to subset those players with height less than 70 inches tall.
- use `filter()` to subset rows of Golden State Warriors ('GSW').
- use `filter()` to subset rows of GSW centers ('C').
- use `filter()` and then `select()`, to subset rows of lakers ('LAL'), and 
then display their names.
- use `filter()` and then `select()`, to display the name and salary, 
of GSW point guards
- find how to select the name, age, and team, of players with more than 
10 years of experience, making 10 million dollars or less.
- find how to select the name, team, height, and weight, of rookie players, 
20 years old, displaying only the first five occurrences (i.e. rows)


-----

## Adding new variables: `mutate()`

Another basic verb is `mutate()` which allows you to add new variables.
Let's create a small data frame for the warriors with three columns: 
`player`, `height`, and `weight`:

```{r}
# creating a small data frame step by step
gsw <- filter(dat, team == 'GSW')
gsw <- select(gsw, player, height, weight)
gsw <- slice(gsw, c(4, 8, 10, 14, 15))
gsw
```

Now, let's use `mutate()` to (temporarily) add a column with the ratio 
`height / weight`:

```{r}
mutate(gsw, height / weight)
```

You can also give a new name, like: `ht_wt = height / weight`:

```{r}
mutate(gsw, ht_wt = height / weight)
```

And add more columns, for example height in meters, and weight in kilograms:

```{r}
mutate(gsw, ht_m = height * 0.0254, wt_kg = weight * 0.4536)
```

In order to permanently change the data, you need to assign the changes to 
an object:

```{r}
gsw2 <- mutate(gsw, ht_m = height * 0.0254, wt_kg = weight * 0.4536)
gsw2
```


## Reordering rows: `arrange()`

The next basic verb of `"dplyr"` is `arrange()` which allows you to reorder 
rows. For example, here's how to arrange the rows of `gsw` by `height`

```{r}
# order rows by height (increasingly)
arrange(gsw, height)
```

By default `arrange()` sorts rows in increasing. To arrande rows in 
descending order you need to use the auxiliary function `desc()`.

```{r}
# order rows by height (decreasingly)
arrange(gsw, desc(height))
```

```{r}
# order rows by height, and then weight
arrange(gsw, height, weight)
```

-----

### Your Turn

- using the data frame `gsw`, add a new variable `product` with the product of 
`height` and `weight`.

- create a new data frame `gsw3`, by adding columns `log_height` and `log_weight`
with the log transformations of `height` and `weight`.

- use the original data frame to `filter()` and `arrange()` those players with height less than 71 inches tall, in increasing order.

- display the name, team, and salary, of the top-5 highest paid players

- display the name, team, and salary, for the top-5 highest paid players

- display the name, team, and points3, of the top 10 three-point players

- create a data frame `gsw_mpg` of GSW players, that contains variables for 
player name, experience, and `min_per_game` (minutes per game), sorted by `min_per_game` (in descending order)


-----

## Summarizing values with `summarise()`

So far we've seen the basic verbs to select (or get access to) the elements in a 
data frame, such as cells, rows, and columns. We've also covered how to add 
new columns, and rearranging rows.

The next verb is `summarise()`. Conceptually, this involves applying a function
on one or more columns, in order to summarize values. This is probably easier 
to understand with one example.

Say you are interested in calculating the average salary of all NBA players.
To do this "a la dplyr" you use `summarise()`, or its synonym function 
`summarize()`:

```{r}
# average salary of NBA players
summarise(dat, avg_salary = mean(salary))
```

Calculating an average like this seems a bit _verbose_, especially
when you can directly use `mean()` like this:

```{r}
mean(dat$salary)
```

So let's make things a bit more interessting. What if you want to calculate some 
summary statistics for `salary`: min, median, mean, and max?

```{r}
# some stats for salary (dplyr)
summarise(
  dat, 
  min = min(salary),
  median = median(salary),
  avg = mean(salary),
  max = max(salary)
)
```

Well, this may still look like not much. You can do the same in base R 
(there are actually better ways to do this):

```{r}
# some stats for salary (base R)
c(min = min(dat$salary), 
  median = median(dat$salary),
  median = mean(dat$salary),
  max = max(dat$salary))
```


## Grouped operations

To actually appreciate the power of `summarise()`, we need to introduce the 
other major basic verb in `"dplyr"`: `group_by()`. This is the function that 
allows you to do perform data aggregations, or _grouped operations_.

Let's see the combination of `summarise()` and `group_by()` to calculate the
average salary by team:

```{r}
# average salary, grouped by team
summarise(
  group_by(dat, team),
  avg_salary = mean(salary)
)
```

Here's a similar example with the average salary by position:

```{r}
# average salary, grouped by position
summarise(
  group_by(dat, position),
  avg_salary = mean(salary)
)
```

Here's a more fancy example: average weight and height, by position, displayed
in desceding order by average height:

```{r}
arrange(
  summarise(
    group_by(dat, position),
    avg_height = mean(height),
    avg_weight = mean(weight)),
  desc(avg_height)
)
```


### Your turn:

- use `summarise()` to get the largest height value.

- use `summarise()` to get the standard deviation of `points3`.

- use `summarise()` and `group_by()` to display the median of three-points,
by team.

- display the average triple points by team, in ascending order, of the 
bottom-5 teams (worst 3pointer teams)

- obtain the mean and standard deviation of `age`, for Power Forwards, with 5 
and 10 years (including) years of experience.

-----
