---
title: "Lab 4: Data Frame Basics"
subtitle: "Stat 133, Fall 2017"
author: "Gaston Sanchez"
output: github_document
urlcolor: blue
---

> ### Learning Objectives:
>
> - Get started with data frames
> - Understand basic operations
> - Understand bracket and dollar notations 

------

## Manipulating Data Frames

The most common format/structure for a data set is in a tabular format 
(with row and columns). When your data is in this shape, most of the times 
you will work with R data frames (or similar structures like a `"matrix"`, 
`"table"`, etc).

Learning how to manipulate data tables is among the most important
_data computing_ basic skills. The traditional way of manipulating data frames 
in R is based on bracket notation, e.g. `dat[ , ]`, to select specific 
rows, columns, or cells. Also, the use of the dollar `$` operator to handle
columns is fundamental. In this part of the lab, you will practice a wide
array of data wrangling tasks with the so-called bracket notation, and the 
dollar operator.

I should say that there are alternative ways for manipulating tables in R
Among the most recent paradigms, there is the __plying__ framework devised
by Hadley Wickham. From his doctoral research, the first tools were available 
in the packages `"plyr"` and `"reshape"`. Nowadays we have the `"reshape2"`
package, and the extremely popular package `"dplyr"` (among other packages).
You will have time to learn more about `"dplyr"` in the next weeks. In the 
meantime, take some time to understand more about the bracket notation.


## Creating data frames

Most of the (raw) data tables you will be working with will already be in 
some data file. However, from time to time you will face the need of creating 
some sort of data table in R. In these situations, you will likely have to 
create such table with a data frame. So let's look at various ways to 
"manually"" create a data frame.

__Option 1__: The primary option to build a data frame is with `data.frame()`. 
You pass a series of vectors (or factors), of the same length, separated by commas.
Each vector (or factor) will become a column. Preferably, give names to each
column, like `col1`, `col2`, and `col3`, in the example below:

```{r create_data_frame1}
# creating a basic data frame
my_table1 <- data.frame(
  col1 = LETTERS[1:5],
  col2 = seq(from = 10, to = 50, by = 10),
  col3 = c(TRUE, TRUE, FALSE, TRUE, FALSE)
)

my_table1
```


__Option 2__: Another way to create data frames is with a `list` (of vectors or 
factors) converted to a data.frame with `data.frame()`. 

```{r create_data_frame2}
# creating a basic data frame
my_list <- list(
  col1 = LETTERS[1:5],
  col2 = seq(from = 10, to = 50, by = 10),
  col3 = c(TRUE, TRUE, FALSE, TRUE, FALSE)
)

my_table2 <- data.frame(my_list)

my_table2
```

Remember that a `data.frame` is nothing more than a `list`. So as long as the 
elements in the list (vectors or factors) are of the same length, we can simply
convert the list into a data frame.

By default, `data.frame()` converts character vectors into factors. You can 
check that by exmining the structure of the data frame with `str()`:

```{r}
str(my_table2)
```

To prevent `data.frame()` from converting strings into factors, you must use
the argument `stringsAsFactors = FALSE`

```{r}
# strings as strings (not as factors)
my_table3 <- data.frame(
  col1 = LETTERS[1:5],
  col2 = seq(from = 10, to = 50, by = 10),
  col3 = c(TRUE, TRUE, FALSE, TRUE, FALSE),
  stringsAsFactors = FALSE
)

str(my_table3)
```


### Your turn

Here's a table with the starting lineup of the Golden State Warriors:

| Player   | Position | Salary     | Points | PPG  | Rookie |
|----------|----------|------------|--------|------|--------|
| Thompson |    SG    | 16,663,575 |  1742  | 22.3 | FALSE  |
| Curry    |    PG    | 12,112,359 |  1999  | 25.3 | FALSE  |
| Green    |    PF    | 15,330,435 |   776  | 10.2 | FALSE  |
| Durant   |    SF    | 26,540,100 |  1555  | 25.1 | FALSE  |
| Pachulia |    C     |  2,898,000 |   426  |  6.1 | FALSE  |

- Start by creating vectors for each of the columns.
- Create a first data frame with `data.frame()`.
- Check that this data frame is of class `"data.frame"`, and that it is also a list.
- Create another data frame by first starting with a `list()`, and then passing 
the list to `data.frame()`.
- What would you do to obtain a data frame such that when you check its structure `str()` the variables are:
    - _Player_ as character
    - _Position_ as factor
    - _Salary_ as numeric or real (ignore the commas)
    - _Points_ as integer
    - _PPG_ as numeric or real
    - _Rookie_ as logical
- Find out how to use the _column binding_ function `cbind()` to create a 
tabular object (see what is the class of this object) with your vectors.
- How could you convert the object in the previous step into a data frame?


## Basic Operations

Now that you have seen some ways to create data frames, let's discuss a number
of basic manipulations of data frames. I will show some examples and then
you'll have the chance to put in practice the following operations:

- Selecting table elements
    + select a given cell
    + select a set of cells
    + select a given row
    + select a set of rows
    + select a given column
    + select a set of columns
- Adding a new column
- Deleting a new column
- Renaming a column
- Moving a column
- Transforming a column


```{r echo = FALSE}
tbl <- data.frame(
  player = c('Thompson', 'Curry', 'Green', 'Durant', 'Pachulia'),
  position = c('SG', 'PG', 'PF', 'SF', 'C'),
  salary = c(16663575, 12112359, 15330435, 26540100, 2898000),
  points = c(1742, 1999, 776, 1555, 426),
  ppg = c(22.3, 25.3, 10.2, 25.1, 6.1),
  rookie = rep(FALSE, 5),
  stringsAsFactors = FALSE
)
```

Let's say you have a data frame `tbl` with the lineup of the Golden State Warriors:

```{r, echo = FALSE, comment = ""}
tbl
```


#### Selecting elements

The data frame `tbl` is a 2-dimensional object: the 1st dimension corresponds
to the rows, while the 2nd dimension corresponds to the columns.
Because `tbl` has two dimensions, the bracket notation involves 
working with the data frame in this form: `tbl[ , ]`.
In other words, you have to specify values inside the 
brackets for the 1st index, and the 2nd index: `tbl[index1, index2]`.

```{r}
# select value in row 1 and column 1
tbl[1,1]

# select value in row 2 and column 5
tbl[2,5]

# select values in these cells
tbl[1:3,3:5]
```


If no value is specified for `index1` then all rows are included. Likewise,
if no value is specified for `index2` then all columns are included.

```{r}
# selecting first row
tbl[1, ]

# selecting third row
tbl[3, ]

# selecting second column
tbl[ ,2]

# selecting columns 3 to 5
tbl[ ,3:5]
```


#### Adding a column

Perhaps the simplest way to add a column is with the dollar operator `$`.
You just need to give a name for the new column, and assign a vector (or factor):

```{r}
# adding a column
tbl$new_column <- c('a', 'e', 'i', 'o', 'u')
tbl
```

Another way to add a column is with the _column binding_ function `cbind()`:

```{r}
# vector of weights
weight <- c(215, 190, 230, 240, 270)

# adding weights to tbl
tbl <- cbind(tbl, weight)
tbl
```


#### Deleting a column

The inverse operation of adding a column consists of __deleting__ a column.
This is possible with the `$` dollar operator. For instance, say you want to 
remove the column `new_column`. Use the `$` operator to select this column,
and assign it the value `NULL` (think of this as _NULLifying_ a column):

```{r}
# deleting a column
tbl$new_column <- NULL
tbl
```


#### Renaming a column

What if you want to rename a column? There are various options to do this.
One way is by changing the column names attribute:

```{r}
# attributes
attributes(tbl)
```

which is more commonly accessed with the `names()` function:

```{r}
# column names
names(tbl)
```

Notice that `tbl` has a list of attributes. The element `names` is the vector
of column names.

You can directly modified the vector of `names`; for example let's change
`rookie` to `rooky`:

```{r}
# changing `rookie` to `rooky`
attributes(tbl)$names[6] <- "rooky"

# display column names
names(tbl)
```

By the way: this way of changing the name of a variable is very low level, and probably
unfamiliar to most useRs.


#### Moving a column

A more challenging operation is when you want to move a column to a different 
position. What if you want to move `salary` to the last position (last column)?
One option is to create a vector of column names in the desired order, and then
use this vector (for the index of columns) to reassign the data frame like this:

```{r}
reordered_names <- c("player", "position", "points", "ppg", "rooky", "weight", "salary")

# moving salary at the end
tbl <- tbl[ ,reordered_names]
tbl
```


#### Transforming a column

A more common operation than deleting or moving a column, is to transform the 
values in a column. THis can be easily accomplished with the `$` operator.
For instance, let's say that we want to transform `salary` from dollars to
millions of dollars:

```{r}
# salary in millions of dollars
tbl$salary <- tbl$salary / 1000000
tbl
```

Likewise, instead of using the `$` operator, you can refer to the column with
bracket notation. Here's how to transform weight from pounds to kilograms
(1 pound = 0.453592 kilograms):

```{r}
# weight in kilograms
tbl[ ,"weight"] <- tbl[ ,"weight"] * 0.453592
tbl
```

There is also the `transform()` function which transform values _interactively_,
that is, temporarily:

```{r}
# transform weight to inches
transform(tbl, weight = weight / 0.453592)
```

`transform()` does its job of modifying the values of `weight` but only 
temporarily; if you inspect `tbl` you'll see what this means:

```{r}
# did weight really change?
tbl
```

To make the changes permanent with `transform()`, you need to reaassign them
to the data frame:

```{r}
# transform weight to inches (permanently)
tbl <- transform(tbl, weight = weight / 0.453592)
tbl
```


-----

## NBA Players Data

```{r nba, echo = FALSE}
dat <- read.csv('../data/nba2017-players.csv', stringsAsFactors = FALSE)
```

The data file for this lab is `nba2017-players.csv`, located in the `data/` 
folder of the course github repository. This file 
contains `r ncol(dat)` variables measured on `r nrow(dat)` players.
Download a copy of the csv file to your computer. 

```r
# download csv file into your working directory
csv <- "https://github.com/ucb-stat133/stat133-fall-2017/raw/master/data/nba2017-players.csv"
download.file(url = csv, destfile = 'nba2017-players.csv')
```

To import the data in R you can use the function `read.csv()`:

```{r read, eval = FALSE}
dat <- read.csv('nba2017-players.csv', stringsAsFactors = FALSE)
```

Notice that I'm specifying the argument `stringsAsFactors = FALSE` to avoid 
the conversion of characters into R factors. Why do you think this is a good 
practice?

All the default reading table functions generate a data frame. Typically, 
everytime I read a new data set which I'm not familiar with, or a data set 
that I haven't worked on in a long time, I always like to call a couple of 
functions to inspect its contents:

- `dim()`
- `head()`
- `tail()`
- `str()`
- `summary()`

A first check-up is to examine the dimensions of the data frame with `dim()`:

```{r}
# dimensions (# of rows, # of columns)
dim(dat)
```

If you know in advanced how many rows and columns are in the data table, this 
is a good way to make sure that R was able to read all the records.

Then, depending on the size of the data, you may want to take a peek 
at its contents with `head()` or `tail()`, just to get an idea of what the data 
looks like:

```{r}
# display first few rows
head(dat)
```

For a more detailed description of how R is treating the data type in each 
column, you should use the structure function `str()`.

```{r}
# check the structure
str(dat, vec.len = 1)
```

This function `str()` displays the dimensions of the 
data frame, and then a list with the name of all the 
variables, and their data types (e.g. `chr` character, `num` real, etc). 
The argument `vec.len = 1` indicates that just the first element in each 
column should be displayed.

When working with data frames, remember to always take some time inspecting 
the contents, and checking how R is handling the data types. It is in these 
early stages of data exploration that you can catch potential issues in order 
to avoid disastrous consequences or bugs in subsequent stages.


-----


### Your turn:

Use bracket notation, the dollar operator, as well as concepts of 
logical subsetting and indexing to:

- Display the last 5 rows of the data.

- Display those rows associated to players having height less than 70 inches tall.

- Of those players that are centers (position `C`), display their names and salaries.

- Create a data frame `durant` with Kevin Durant's information (i.e. row).

- Create a data frame `ucla` with the data of players from college UCLA.

- Create a data frame `rookies` with those players with 0 years of experience. 

- Create a data frame `rookie_centers` with the data of Center rookie players. 

- Create a data frame for players with more than 50 games and more than 100 minutes played.

- What's the largest height value?

- What's the minimum height value?

- What's the overall average height?

- Who is the tallest player?

- Who is the shortest player?

- Which are the unique teams?

- How many different teams?

- Who is the oldest player?

- What is the median salary?

- How many players scored 4 points or less?

- Who are those players (who scored 4 points or less)?

- Who is the player with 0 points?

- Are there any players from "University of California, Berkeley"? If so how many and who are they?

- Are there any players from "University of California, Los Angeles"? If so how many and who are they?

- Are there any players with weight greater than 260 pounds? If so how many and who are they?

- How many players did not attend a college in the US?

- Who is the player with the maximum rate of points per minute?

- Who is the player with the maximum rate of three-points per minute?

- Who is the player with the maximum rate of two-points per minute?

- Who is the player with the maximum rate of one-points (free-throws) per minute?


### Sorting

Use `sort()`, `order()`, `rev()`, together with data frame manipulations, 
to:

- Obtain height values in increasing order

- Obtain weight values in decreasing order

- Sort data by height in increasing order

- Sort data by weight in decreasing order

- Sort data by gender in increasing order

- Sort data by species in increasing order


## Group By

Group-by operations are very common in data analytics. Without dedicated 
functions, these operations tend to be very hard (labor intensive). 

__Quick try__: Using just bracket notation, try to create a data frame with 
two columns: the team name, and the team payroll (addition of all player salaries).

So what functions can you use in R to perform group by operations? In base R, 
the main function for group-by operations is `aggregate()`.

Here's an example using `aggregate()` to get the median salary, grouped by team:

```{r}
aggregate(dat$salary, by = list(dat$team), FUN = median)
```

The same example above can also be obtained with `aggreagte()` using formula 
notation like this:

```{r}
aggregate(salary ~ team, data = dat, FUN = median)
```


Here's another example using `aggregate()` to get the average height and 
average weight, grouped by position:

```{r}
aggregate(dat[ ,c('height', 'weight')], by = list(dat$position), FUN = mean)
```

The same example above can also be obtained with `aggreagte()` using formula 
notation like this:

```{r}
aggregate(. ~ position, data = dat[ ,c('position', 'height', 'weight')],
          FUN = mean)
```

### Your turn

- Create a data frame with the average height, average weight, and average 
age, grouped by position

- Create a data frame with the average height, average weight, and average 
age, grouped by team

- Create a data frame with the average height, average weight, and average 
age, grouped by team and position.

- Difficult: Create a data frame with the minimum salary, median salary,
mean salary, and maximum salary, grouped by team and position.


-----
