---
title: "Lab 6: Simple Conditionals"
subtitle: "Stat 133, Fall 2017"
author: "Gaston Sanchez"
output: github_document
fontsize: 11pt
urlcolor: blue
---

> ### Learning Objectives
>
> - Get into the habit of writing simple functions
> - Making sure your functions work
> - Understand the concept of _vectorization_

------


## If Conditionals

Write R code that will "squish" a number into the interval [0, 100], so that a 
number less than 0 is replaced by 0 and a number greater than 100 is replaced 
by 100.

```{r}
z <- 100*pi
# Fill in the following if-else statements. You may (or may not) 
# have to add or subtract else if or else statements.
if (TRUE) { # Replace TRUE with a condition.
  
} else if (TRUE) { # Replace TRUE with a condition.
  
} else {
  
}
```

### Multiple If's

A common situation involves working with multiple conditions at the same time. 
You can chain multiple if-else statements:

```{r}
y <- 1 # Change this value!

if (y > 0) {
  print("positive")
} else if (y < 0) {
  print("negative")
} else {
  print("zero?")
}
```


## Even number

Write a function `is_even()` that determines whether a number is even 
(i.e. multiple of 2). If the input number is even, the output should be 
`TRUE`. If the input number is odd, the output should be `FALSE`. 
If the input is not a number, the output should be `NA` 
```{r echo = FALSE}
is_even <- function(x) {
  if (is.numeric(x)) {
    return(x %% 2 == 0)
  } else {
    return(NA)
  }
}
```

For example:
```{r}
# even number
is_even(10)

# odd number
is_even(33)

# not a number
is_even('a')
```


## Odd number

Use your function `is_even()` to write a function `is_odd()` that determines 
if a number is odd (i.e. not a multiple of 2). If a number is odd, the output 
should be `TRUE`; if a number is even the output should be `FALSE`; if the 
input is not a number the output should be `NA`
```{r, echo = FALSE}
is_odd <- function(x) {
  !is_even(x)
}
```

For example:
```{r}
# odd number
is_odd(1)

# even number
is_odd(4)

# not a number
is_odd('a')
```


## Switch

Working with multiple chained if's becomes cumbersome. Consider the following
example that uses several if's to convert a day of the week into a number:
```{r}
# Convert the day of the week into a number.
day <- "Tuesday" # Change this value!

if (day == 'Sunday') {
  num_day <- 1
} else {
  if (day == "Monday") {
    num_day <- 2
  } else {
    if (day == "Tuesday") {
      num_day <- 3
    } else {
      if (day == "Wednesday") {
        num_day <- 4
      } else {
        if (day == "Thursday") {
          num_day <- 5
        } else {
          if (day == "Friday") {
            num_day <- 6
          } else {
            if (day == "Saturday") {
              num_day <- 7
            }
          }
        }
      }
    }
  }
}


num_day
```

Working with several nested if's like in the example above can be a nigthmare.

In R, you can get rid of many of the braces like this:

```{r}
# Convert the day of the week into a number.
day <- "Tuesday" # Change this value!

if (day == 'Sunday') {
  num_day <- 1
} else if (day == "Monday") {
  num_day <- 2
} else if (day == "Tuesday") {
  num_day <- 3
} else if (day == "Wednesday") {
  num_day <- 4
} else if (day == "Thursday") {
  num_day <- 5
} else if (day == "Friday") {
  num_day <- 6
} else if (day == "Saturday") {
  num_day <- 7
}

num_day
```

But still we have too many if's, and there's a lot of repetition in the code.
If you find yourself using many if-else statements with identical structure for 
slightly different cases, you may want to consider a __switch__ statement 
instead:

```{r}
# Convert the day of the week into a number.
day <- "Tuesday" # Change this value!

switch(day, # The expression to be evaluated.
  Sunday = 1,
  Monday = 2,
  Tuesday = 3,
  Wednesday = 4,
  Thursday = 5,
  Friday = 6,
  Saturday = 7,
  NA) # an (optional) default value if there are no matches
```

Switch statements can also accept integer arguments, which will act as indices 
to choose a corresponding element:

```{r}
# Convert a number into a day of the week.
day_num <- 3 # Change this value!

switch(day_num,
  "Sunday",
  "Monday",
  "Tuesday",
  "Wednesday",
  "Thursday",
  "Friday",
  "Saturday")
```

## Converting Miles to other units

The table below shows the different formulas for converting miles (mi) into other scales:

| Units   | Formula         |
| ------- | --------------- |
| Inches  | mi x 63360      |
| Feet    | mi x 5280       |
| Yards   | mi x 1760       |
| Meters  | mi / 0.00062137 |
| Kms     | mi / 0.62137    |

Write the following five functions for each type of conversion. Each function 
must take one argument `x` with default value: `x = 1`.

- `miles2inches()`
- `miles2feet()`
- `miles2yards()`
- `miles2meters()`
- `miles2kms()`

```{r, echo = FALSE}
miles2inches <- function(x = 1) {
  x * 63360
}

miles2feet <- function(x = 1) {
  x * 5280
}

miles2yards <- function(x = 1) {
  x * 1760
}

miles2meters <- function(x = 1) {
  x / 0.00062137
}

miles2kms <- function(x = 1) {
  x / 0.62137
}
```

For example: 
```{r}
miles2inches(2)
miles2feet(2)
miles2yards(2)
miles2meters(2)
miles2kms(2)
```

-----

## Using switch()

Create a function `convert()` that converts miles into the specified units. 
Use `switch()` and the previously defined functions---`miles2inches()`, 
`miles2feet()`, ..., `miles2kms`---to define `convert()`. Use two arguments: 
`x` and `to`, like this:
```{r, eval=FALSE}
convert(40, to = "in")
```

By default, `to = "km"`, but it can take values such as `"in"`, `"ft"`, `"yd"`, or `"m"`.
```{r echo=FALSE}
convert <- function(x, to = "km") {
  switch(to,
         "in" = miles2inches(x),
         "ft" = miles2feet(x),
         "yd" = miles2yards(x),
         "m" = miles2meters(x),
         "km" = miles2kms(x))
}
```

For instance:
```{r eval=FALSE}
convert(3, "in")
convert(3, "ft")
convert(3, "yd")
convert(3, "m")
convert(3, "km")
```


## One more swtich

Write a switch statement to determine if a given lowercase letter is a vowel. 
Assume that "y" is not a vowel.
```{r}
letter <- "e" # Change this value!

switch(letter,
  a = "vowel") # Modify as necessary.
```

