---
title: "Lab 11: Manipulating strings, part 2"
subtitle: "Stat 133, Fall 2017"
author: "Gaston Sanchez"
output: github_document
---

> ### Learning Objectives
>
> - String manipulation
> - Base R functions
> - R package `"stringr"`
> - A bit of data cleaning

------

## Motivation

So far we've been working with data sets that have been already cleaned, and 
can be imported in R ready to be analyzed.

Today we are going to start dealing with "messy" datasets. Most real life data 
sets require a pre-processing phase, and most of the time in any data analysis
project will be spent on getting the data in the right shape. So it is 
extremely important that you gain skills to help you clean raw data.


### Mobile Food Schedule Data

The data set is on the github repo:
[https://raw.githubusercontent.com/ucb-stat133/stat133-fall-2017/master/data/mobile-food-sf.csv](https://raw.githubusercontent.com/ucb-stat133/stat133-fall-2017/master/data/mobile-food-sf.csv)

The original source comes from the SF Open Data website:
[https://data.sfgov.org/Economy-and-Community/Mobile-Food-Schedule/jjew-r69b](https://data.sfgov.org/Economy-and-Community/Mobile-Food-Schedule/jjew-r69b)

Download a copy of the file to your working directory:

```{r eval = FALSE}
github <- "https://raw.githubusercontent.com/ucb-stat133/stat133-fall-2017/master/"
datafile <- "data/mobile-food-sf.csv"
download.file(paste0(github, datafile), destfile = "mobile-food-sf.csv")
```

Once you've downloaded the data file, you can read it in R:

```{r eval = FALSE}
dat <- read.csv('mobile-food-sf.csv', stringsAsFactors = FALSE)
```

```{r echo = FALSE}
dat <- read.csv('../data/mobile-food-sf.csv', stringsAsFactors = FALSE)
```

The variables are:

- `DayOfWeekStr`
- `starttime`
- `endtime`
- `PermitLocation`
- `optionaltext`
- `ColdTruck`
- `Applicant`
- `Location`


### Package `"stringr"`

R has a set of built-in functions for manipulating strings. But there is also
the package `"stringr"` that provides more functionality and a more user 
friendly syntax. 

```{r}
# install.packages("stringr")
library(stringr)
```


## Changing Times

Let's begin processing the values in column `starttime`. The goal is to obtain 
new times in 24 hr format. For example, a starting time of `10AM` will be 
transformed to `10:00`. Likewise, a starting time of `1PM` will be
transformed to `13:00`.

We are going to be manipulating character strings. Hence, I recommend to start working on a small subset of values. Figure out how to get the
answers working on this subset, and then generalize to the entire data set.

Consider the first starting time that has a value of `10AM`. To get a better feeling of string manipulation, let's create a toy string with this value:

```{r}
# toy string
time1 <- '10AM'
```


### Function `str_sub()`

To get the time and period values, you can use `str_sub()`:

```{r}
# hour
str_sub(time1, start = 1, end = 2)

# period
str_sub(time1, start = 3, end = 4)
```

__Your turn__: What about times where the hour has just one digit? For example: `9AM`, or `8AM`? Create the following vector `times` and try to subset the hour and the periods with `str_sub()` 

```{r}
times <- c('12PM', '10AM', '9AM', '8AM')

# subset time


# subset period


#
```

The nice thing about `str_sub()` is that it allows you to specify negative values for the `start` and `end` positions. Run the command below and see what happens:

```{r eval = FALSE}
# period
str_sub(times, start = -2)
```


### Function `str_replace()`

The tricky part with the vector `times` is the extraction of the hour. One solution is to "remove" the
characters `AM` or `PM` from each time. You can do this with the substitution
function `str_replace()`:

```{r}
str_replace(times, pattern = 'AM|PM', replacement = '')
```


### Your Turn

So far you've managed to get the hour value and the period. Notice that 
you still need to convert the hours as numeric vectors:

```{r}
hours <- as.numeric(str_replace(times, pattern = 'AM|PM', replacement = ''))
periods <- str_sub(times, start = -2)
```

- Transform the hours into 24 hr values. Write R code to
create a vector `start24` that contains the hour in 24hr scale.

```{r}
# your vector start24

```


- Add two columns `start` and `end` to the data frame `dat`, containing
the starting and ending hour respectively (columns must be `"numeric"`).

```{r}
# add 'start' hours


# add 'end' hours

```


- With the starting and ending hours, calculate the duration, and add one
more column `duration` to the data frame `dat`:

```{r}
# add 'duration'

```


-----

## Lat and Long Coordinates

Another interesting column in the data is `Location`. If you look at this 
column, you will see values like the string below `loc1`

```{r}
loc1 <- "(37.7651967350509,-122.416451692902)"
```

The goal is to split `Location` into latitude and longitude.

First we need to remove the parenthesis. The issue here is that the characters
`(` and `)` have special meanings; recall they are metacharacters. So you need to __escape__ them by 
pre-appending two backslashes: `\\(` and `\\)`

```{r}
# "remove" opening parenthesis 
str_replace(loc1, pattern = '\\(', replacement = '')

# "remove" closing parenthesis
str_replace(loc1, pattern = '\\)', replacement = '')
```

You can also combine both patterns in one single call. But be careful:

```{r}
str_replace(loc1, pattern = '\\(|\\)', replacement = '')
```

`str_replace()` replaces only the first occurrence of `(` or `)`. However, the
location values contain both opening and closing parentheses. To replace them
all, you have to use `str_replace_all()`

```{r}
str_replace_all(loc1, pattern = '\\(|\\)', replacement = '')
```

Now we need to get rid of the comma `,`. You could replace it with an empty
string, but then you will end up with one long string like this:

```{r}
lat_lon <- str_replace_all(loc1, pattern = '\\(|\\)', replacement = '')

str_replace(lat_long, pattern = ',', replacement = '')
```

Instead of replacing the comma, what we need to use is `str_split()`

```{r}
# string split in stringr
str_split(lat_lon, pattern = ',')
```

Notice that `str_split()` returns a list.


### Manipulating more location values

Let's define a vector with more location values, so we can start generalizing
our code:

```{r}
locs <- c(
  "(37.7651967350509,-122.416451692902)",
  "(37.7907890558203,-122.402273431333)",
  "(37.7111991003088,-122.394693339395)",
  "(37.7773000262759,-122.394812784799)",
  NA
)
```

- create a list `lat_lon` containing the latitude and the longitude values of `locs`

Assuming that you have `lat_long`, to retrieve the latitude and longitude values, you can use the `lapply()` 
function, and then specify an _anonymous_ function to get the first element 
(for the latitude) and the second element (for the longitude):

```{r eval = FALSE}
lat <- lapply(lat_lon, function(x) x[1])
long <- lapply(lat_lon, function(x) x[2])
```

Finally, to convert from list to a vector, use `unlist()`

```{r eval = FALSE}
latitute <- as.numeric(unlist(lat))
longitude <- as.numeric(unlist(lon))
```


### Your Turn

Add two more columns: `lat` and `lon` to `dat`

```{r}
# your code

```


-----

## Exporting the cleaned data

Now that you've pre-processed some of the columns in `dat`, it's time to create
a clean version of the data set.

Select the columns:

- `DayOfWeekStr`
- `start`
- `end`
- `duration`
- `PermitLocation`
- `optionaltext`
- `ColdTruck`
- `Applicant`
- `lat`
- `lon`

Save the data with the selected columns as a csv file called 
`food-mobile-clean.csv` (save it in your working directory)

```{r}
# export the clean data

```

-----
